---
title: "Appendix"
author: "ibdj"
format:
  pdf:
    fig-width: 6.5  # Approx 16.5cm, slightly wider than text to be safe
    out-width: 100%
    fig-align: center
    toc: true
editor: visual
---

# Current climate conditions

Data is available 2007 – 2022. 2008 is the first whole year of temperature data. Data available from GEM. (2020). ClimateBasis Nuuk—Temperature—Air temperature \@ 200 cm—30min average (°C) (Version 1.0) \[Dataset\]. Greenland Ecosystem Monitoring. <https://doi.org/10.17897/PGN3-7597>

### Tempterature

```{r}
#| label: importing-packages
#| output: FALSE
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

#install.packages("rmarkdown","pacman","tidyverse","yaml", dependencies = TRUE)
library(pacman)
pacman::p_load(ggplot2, Hmisc, tidyverse, janitor, lubridate, glue, hms,knitr, broom, ggpmisc, yaml,readxl,zoo,emmeans,lme4,multcomp)

opts_chunk$set( size="small")

# colors
sleet <- rgb(110/255, 107/255, 196/255, 1)
rain <- rgb(0/255, 0/255, 165/255, 1)
warm <- rgb(165/255, 0/255, 0/255, 1)
```

```{r}
#| label: importing-temp-data
#| echo: FALSE


data_temp <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/temperature/data/View_ClimateBasis_Nuuk_Data_Temperature_Air_temperature_@_200_cm__30min_average__DegreesC220420240157436123.csv", delim = "\t", escape_double = FALSE,col_types = cols(Date = col_datetime(format = "%Y-%m-%d"), Time = col_time(format = "%H:%M:%S")), trim_ws = TRUE) |> 
  janitor::clean_names() |> 
  drop_na() |>  
  filter(at_c != -9999) |>  
   mutate(year = year(date),
         month = month(date), 
         doy = yday(date),
         month_string = month.name[month])
```

```{r}
#| label: temp-mean-monthly
#| echo: false

data_temp_monthly_mean <- data_temp |> 
  group_by(month) |> 
  dplyr::reframe(mean = round(mean(at_c),2),
            sd = round(sd(at_c),2)) |> 
  mutate(month_string = month.name[month]) |> 
  arrange(desc(mean)) |> 
  mutate(temp_rank = row_number())

year_mean <- round(mean(data_temp$at_c),4)

data_temp_annual_mean <- data_temp |> 
  group_by(year) |> 
  filter(year != 2007) |> #excluding 2007 because it is only the months of oct, nov, dec
  dplyr::reframe(mean = mean(at_c),
            sd = sd(at_c))

```
#### Monthly mean

```{r}
#| label: fig-month-mean-temp
#| fig-scap: "Mean air temperature (°C) pr month"
#| fig-cap: !expr glue("Mean temperature pr month. Mean air temperature (°C) pr month, based on data from {min(data_temp$year)} to {max(data_temp$year)}. The dashed line represent yearly mean of {round(year_mean,3)}°C.")
#| echo: false
#| message: false
#| warning: false

ggplot(data_temp_monthly_mean, aes(x = month, y = mean))+
  geom_line(size = 0.25, linetype="solid", alpha = 0.2)+
  geom_smooth(color = "gray", se = F,  alpha = 0.1)+
  geom_point(size = 3, alpha = 0.7, color = "#076834") +
    scale_x_continuous(breaks=seq(1, 12, by = 1))+
  geom_hline(yintercept=year_mean, linetype="dotted", color = "#076834", size=0.5)+
  labs(x = "Month", y = "Mean temperature (°C)", title = "")+
  theme_minimal()+
  theme(axis.title.x = element_text(size = 8), axis.title.y = element_text(size = 8))
```

#### Yearly mean

```{r}
#| label: fig-annual-mean-temp
#| fig-scap: "Mean air temperature pr year (°C)"
#| fig-cap: !expr glue("Annual mean temperature, Kangerluassunguaq. Mean air temperature pr year (°C) from {min(data_temp_annual_mean$year)} to {max(data_temp_annual_mean$year)}. Data from 2007 is excluded because data is only from October, November and December. Gray dotted line indicate over all mean of {year_mean}°C. Dashed line is trend line.")
#| message: false
#| warning: false
#| echo: false

ggplot(data_temp_annual_mean, aes(x = year, y = mean))+
  geom_line(size = 0.4, color = "gray", linetype="solid", alpha = 0.5)+
  geom_smooth(method = lm, color = "#076834", se = F, linetype = "dashed")+
  geom_point(size = 3, alpha = 0.7, color = "#076834") +
    scale_x_continuous(breaks=seq(min(data_temp_annual_mean$year), max(data_temp_annual_mean$year), by = 1))+
  geom_hline(yintercept=year_mean, linetype="dotted", color = "darkgray", size=0.7)+
  labs(x = "Year", y = "Mean temperature (°C)", title = "")+
  stat_poly_eq(
      aes(label = paste(..eq.label.., ..rr.label.., ..p.value.label.., sep = "~~~")),
      formula = y ~ x,
      parse = TRUE
    )+
  theme_minimal()+
  theme(axis.title.x = element_text(size = 8), axis.title.y = element_text(size = 8))
```
#### Five year intervals
## Precipitation
### Monthly mean
### Yearly mean
### Five year intervals
# Historic climate data
## Temperature
## Precipitation
# Statistical analysis
## Diversity
### Richness
### Evenness
### Shannon
### Turnover
## Community composition
### Within vegetation composition
### Between vegetation composition
## Shrubification
### Functional groups
### Specific species

