---
title: "Appendix"
author: "Jacobsen et al 2026"
format:
  pdf:
    fig-width: 6.5  # Approx 16.5cm, slightly wider than text to be safe
    out-width: 100%
    fig-align: center
    toc: true
    lof: true
    fig-pos: "H"
    include-in-header:
      text: |
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \fancyhf{}
        \lhead{Appendix: NERO line paper}
        \rhead{\thepage}
        \renewcommand{\headrulewidth}{0.4pt}
        \setlength{\headheight}{15pt}
        \fancypagestyle{plain}{
          \fancyhf{}
          \lhead{Appendix: NERO line paper}
          \rhead{\thepage}
          \renewcommand{\headrulewidth}{0.4pt}
        }
editor: visual
---

# Current climate conditions

Data is available 2007 – 2022. 2008 is the first whole year of temperature data. Data available from GEM. (2020). ClimateBasis Nuuk—Temperature—Air temperature \@ 200 cm—30min average (°C) (Version 1.0) \[Dataset\]. Greenland Ecosystem Monitoring. <https://doi.org/10.17897/PGN3-7597>

## Tempterature

```{r}
#| label: importing-packages
#| output: false
#| echo: false
#| message: false
#| warning: false

#install.packages("rmarkdown","pacman","tidyverse","yaml", dependencies = TRUE)
library(pacman)
pacman::p_load(ggplot2, Hmisc, tidyverse, janitor, lubridate, glue, hms,knitr, broom, ggpmisc, yaml,readxl,zoo,emmeans,lme4,multcomp, vegan,lmerTest)

opts_chunk$set( size="small")

# colors
sleet <- rgb(110/255, 107/255, 196/255, 1)
rain <- rgb(0/255, 0/255, 165/255, 1)
warm <- rgb(165/255, 0/255, 0/255, 1)
```

```{r}
#| label: importing-temp-data
#| echo: FALSE


data_temp <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/temperature/data/View_ClimateBasis_Nuuk_Data_Temperature_Air_temperature_@_200_cm__30min_average__DegreesC220420240157436123.csv", delim = "\t", escape_double = FALSE,col_types = cols(Date = col_datetime(format = "%Y-%m-%d"), Time = col_time(format = "%H:%M:%S")), trim_ws = TRUE) |> 
  janitor::clean_names() |> 
  drop_na() |>  
  filter(at_c != -9999) |>  
   mutate(year = year(date),
         month = month(date), 
         doy = yday(date),
         month_string = month.name[month])
```

```{r}
#| label: temp-mean-monthly
#| echo: false

data_temp_monthly_mean <- data_temp |> 
  group_by(month) |> 
  dplyr::reframe(mean = round(mean(at_c),2),
            sd = round(sd(at_c),2)) |> 
  mutate(month_string = month.name[month]) |> 
  arrange(desc(mean)) |> 
  mutate(temp_rank = row_number())

year_mean <- round(mean(data_temp$at_c),4)

data_temp_annual_mean <- data_temp |> 
  group_by(year) |> 
  filter(year != 2007) |> #excluding 2007 because it is only the months of oct, nov, dec
  dplyr::reframe(mean = mean(at_c),
            sd = sd(at_c))

```

#### Monthly mean

```{r}
#| label: fig-month-mean-temp
#| fig-scap: "Mean air temperature (°C) pr month"
#| fig-cap: !expr glue("Mean temperature pr month. Mean air temperature (°C) pr month, based on data from {min(data_temp$year)} to {max(data_temp$year)}. The dashed line represent yearly mean of {round(year_mean,3)}°C.")
#| echo: false
#| message: false
#| warning: false

ggplot(data_temp_monthly_mean, aes(x = month, y = mean))+
  geom_line(size = 0.25, linetype="solid", alpha = 0.2)+
  geom_smooth(color = "gray", se = F,  alpha = 0.1)+
  geom_point(size = 3, alpha = 0.7, color = "#076834") +
    scale_x_continuous(breaks=seq(1, 12, by = 1))+
  geom_hline(yintercept=year_mean, linetype="dotted", color = "#076834", size=0.5)+
  labs(x = "Month", y = "Mean temperature (°C)", title = "")+
  theme_minimal()+
  theme(axis.title.x = element_text(size = 8), axis.title.y = element_text(size = 8))
```

#### Yearly mean

```{r}
#| label: model-annual-mean-temp
#| 

model_temp_annual_mean <- lm(mean ~ year, data = data_temp_annual_mean)
summary(model_temp_annual_mean)

data_temp_annual_mean$pred <- predict(model_temp_annual_mean)

```

```{r}
#| label: fig-annual-mean-temp
#| fig-scap: "Mean air temperature pr year (°C)"
#| fig-cap: !expr glue("Mean annual air temperature pr year (°C) from {min(data_temp_annual_mean$year)} to {max(data_temp_annual_mean$year)} in Kangerluassunguaq. Data from 2007 is excluded because data is only from October, November and December. Gray dotted line indicate over all mean of {year_mean}°C. Dashed line is model prediction.")
#| message: false
#| warning: false
#| echo: false

ggplot(data_temp_annual_mean, aes(x = year, y = mean))+
  geom_line(size = 0.4, color = "gray", linetype="solid", alpha = 0.5)+
  #geom_smooth(method = lm, color = "#076834", se = F, linetype = "dashed")+
  geom_line(
    aes(y = pred),
    color = "#076834",
    linetype = "dashed",
  ) +
  geom_point(size = 3, alpha = 0.7, color = "#076834") +
    scale_x_continuous(breaks=seq(min(data_temp_annual_mean$year), max(data_temp_annual_mean$year), by = 1))+
  geom_hline(yintercept=year_mean, linetype="dotted", color = "darkgray", size=0.7)+
  labs(x = "", y = "Mean temperature (°C)", title = "")+
  theme_minimal()+
    theme(  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8),
  axis.text.x  = element_text(size = 7),
  axis.text.y  = element_text(size = 7))
```

#### Five year intervals

```{r}
#| label: data-fiveyear_intervals
#| message: false
#| warning: false
#| echo: false

data_temp_fiveyear <- data_temp |> 
  mutate(
    # choose the anchor year so that 2007–2022 is a bin
    anchor    = 2007,
    seg_start = anchor + floor((year - anchor) / 5) * 5,
    seg_end   = seg_start + 4,
    segment   = paste0(seg_start, "-", seg_end)
  ) |>
  group_by(segment, seg_start, seg_end) |>
  mutate(seg_mean_temp = mean(at_c, na.rm = TRUE))

interval_lines <- data_temp_fiveyear |>
  filter(year > 2007) |> 
  distinct(segment, seg_start, seg_end, seg_mean_temp)

```

```{r}
#| label: data-fiveyear_intervals_test
#| message: false
#| warning: false
#| echo: false

# Fit the linear mixed model
temp_data <- data_temp_fiveyear |> filter(year > 2007)

# Fit the model explicitly
model_temp_fiveyear <- lmer(
  at_c ~ segment + (1 | year), 
  data = temp_data
)

# Get estimated marginal means with cld
emm_temp_fiveyear <- model_temp_fiveyear |>
  emmeans(~ segment) |>
  cld(alpha = 0.05, Letters = letters)

cld_temp <- as.data.frame(emm_temp_fiveyear)

cld_plot <- cld_temp |>
  left_join(interval_lines, by = "segment")

# View the results
print(emm_temp_fiveyear)

```

```{r}
#| label: fig-temp-fiveyear
#| fig-scap: "5 year intervals in temperature means"
#| fig-cap: !expr glue("Five year intervals in temperature means. Shared letters indicate non-significant differences among estimated means according to compact letter display.")
#| message: false
#| warning: false
#| echo: false

yearly_means <- data_temp_fiveyear |>
  group_by(year) |>
  filter(year > 2007) |> 
  summarise(
    mean_temp = mean(at_c, na.rm = TRUE),
    .groups = "drop"
  )

interval_lines <- data_temp_fiveyear |>
  filter(year > 2007) |> 
  distinct(segment, seg_start, seg_end, seg_mean_temp)

# amke the positions of the cld in the plot
cld_plot <- cld_plot |>
  mutate(
    x_mid = (seg_start + seg_end) / 2
  )

# 3. Plot
ggplot() +
  # Base: Yearly means (points)
  geom_hline(yintercept=year_mean, linetype="dotted", color = "darkgray", size=0.7)+
  geom_point(
    data = yearly_means,
    aes(x = year, y = mean_temp),
    size = 3,
    color = "darkgreen"
  ) +
  
  # Layer: Interval lines (horizontal spans)
  geom_segment(
    data = interval_lines,
    aes(
      x = seg_start,
      xend = seg_end,
      y = seg_mean_temp,
      yend = seg_mean_temp
    )
  ) +
  geom_text(
  data = cld_plot,
  aes(
    x = x_mid,
    y = max(yearly_means$mean_temp)*0.8,
    label = .group
  ),
  vjust = -0.8,
  size = 3
)+
  # Labels and Theme
  labs(
    x = "",
    y = "Mean Annual Temperature (°C)",
    title = ""
  ) +
  theme_minimal()+
  scale_x_continuous(breaks=seq(min(yearly_means$year), max(yearly_means$year), by = 1))+
  theme(  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8),
  axis.text.x  = element_text(size = 7),
  axis.text.y  = element_text(size = 7))
```

## Precipitation

Data is available from GEM. (2020). *ClimateBasis Nuuk—Precipitation—Precipitation accumulated (mm)* (Version 1.0) \[Dataset\]. Greenland Ecosystem Monitoring. <https://doi.org/10.17897/SXJ8-WA79>. Downloaded on XXXX.

```{r}
#| label: importing-precip
#| message: false
#| warning: false
#| echo: false

data_precip <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/precipitation/Precipitation accumulated (mm)_10.17897_SXJ8-WA79/Precipitation accumulated (mm)_10.17897_SXJ8-WA79_data.txt", delim = "\t", escape_double = FALSE, col_types = cols(Date = col_date(format = "%Y-%m-%d"), Time = col_time(format = "%H:%M:%S")), trim_ws = TRUE) |> 
  janitor::clean_names() |> 
  drop_na() |>
  filter(pre_mm != -9999) |> 
   mutate(year = year(date),
         month = month(date),
         doy = yday(date),
         month_string = month.name[month])
```

```{r}
#| label: data-precip-mean-calculations
#| message: false
#| warning: false
#| echo: false

data_precip_sum <- data_precip |> 
  group_by(year) |> 
  filter(year != 2007) |> 
  summarise(pre_mm_sum = sum(pre_mm))
data_precip_mean <- data_precip_sum |> 
  summarize(mean = mean(pre_mm_sum))

precip_year_mean <- round(mean(data_precip_mean$mean),2)

data_precip_mean_year <- data_precip  |> 
  group_by(year) |> 
  summarise(yearly_precip = sum(pre_mm, na.rm = TRUE)) |> 
  summarise(mean_precip = mean(yearly_precip, na.rm = TRUE))  |> 
  pull(mean_precip)

data_temp_hour <- data_temp |> 
  mutate(hour = substr(time, 1, 2)) |> 
  group_by(date, hour) |> 
  summarise(at_c = mean(at_c, na.rm = TRUE)) |> 
  mutate(time = hms::hms(hours = as.numeric(hour)))

sleet_lower <- -1
sleet_upper <- 1 
  
  data_pre_tem_joined <- data_precip |> 
  left_join(data_temp_hour, by = c("date" = "date", "time" = "time")) |> 
  drop_na() |> 
  mutate(precip_type = case_when(at_c < sleet_lower ~ "snow",
                              at_c <= sleet_upper ~ "sleet",
                              TRUE ~ "rain"))
  
  data_precip_month_sum <- data_precip |> 
  group_by(month, year) |> 
  summarise(pre_mm_sum = sum(pre_mm))

data_precip_month_mean <- data_precip_month_sum |> 
  group_by(month) |> 
  summarise(pre_mm_mean = round(mean(pre_mm_sum),2),
         sd = sd(pre_mm_mean)) |> 
  mutate(month_string = month.name[month]) |> 
  arrange(desc(pre_mm_mean)) |> 
  mutate(temp_rank = row_number())

precip_month_mean <- round(mean(data_precip_month_mean$pre_mm_mean),2)

data_precip_mean_annual <- data_precip  |> 
  filter(year != 2007) |> 
  group_by(year) |> 
  summarise(yearly_precip = sum(pre_mm, na.rm = TRUE))

```

The mean annual precipitation is `r precip_year_mean` mm.

The months with the most precipitation are `r data_precip_month_mean$month_string[1]` (`r data_precip_month_mean$pre_mm_mean[1]` mm), `r data_precip_month_mean$month_string[2]` (`r data_precip_month_mean$pre_mm_mean[2]` mm), `r data_precip_month_mean$month_string[3]` (`r data_precip_month_mean$pre_mm_mean[3]` mm), which account for `r data_precip_month_mean$pre_mm_mean[1]+data_precip_month_mean$pre_mm_mean[2]+data_precip_month_mean$pre_mm_mean[3]` mm or `r  round((data_precip_month_mean$pre_mm_mean[1]+data_precip_month_mean$pre_mm_mean[2]+data_precip_month_mean$pre_mm_mean[3])/precip_year_mean*100,1)` % of the yearly precipitation.

The months with the least precipitation are `r data_precip_month_mean$month_string[10]` (`r data_precip_month_mean$pre_mm_mean[10]` mm), `r data_precip_month_mean$month_string[11]` (`r data_precip_month_mean$pre_mm_mean[11]` mm), `r data_precip_month_mean$month_string[12]` (`r data_precip_month_mean$pre_mm_mean[12]` mm).

#### Monthly mean

```{r}
#| label: data-yearly-mean-dist_rain-snow
#| echo: false
#| message: false

# joint the temperature precipitation data to distinguish snow and rain. 

data_pre_tem_joined_monthly <- data_pre_tem_joined |> 
  group_by(year,month,precip_type) |> 
  summarise(sum = sum(pre_mm)) |> 
  filter(year != 2007)

data_pre_tem_monthly <- data_pre_tem_joined_monthly |> 
  group_by(month,precip_type) |> 
  summarise(mean = round(mean(sum),2)) 

data_pre_tem_year <- data_pre_tem_joined_monthly |> 
  group_by(year,precip_type) |> 
  summarise(mean = round(sum(sum),2)) 
```

```{r}
#| label: fig-snow-rain-year
#| echo: false
#| message: false
#| warning: false
#| fig-scap: "Mean monthly distribution of types of precipitation (mm)"
#| fig-cap: !expr glue("Mean precipitation type pr month. Mean monthly distribution of types of precipitation (mm). Based on data from {min(data_pre_tem_joined_monthly$year)} to {max(data_pre_tem_joined_monthly$year)}. Sleet is defined as precipitation that fell when temperatures were between {sleet_lower}°C and {sleet_upper}°C.")

ggplot(data_pre_tem_monthly, aes(x = month, y = mean, fill = precip_type)) + 
  geom_bar(stat = "identity")+
  scale_x_continuous(breaks=seq(min(data_pre_tem_monthly$month), max(data_pre_tem_monthly$month), by = 1))+
  labs(x = "Month", y = "Mean precipitation (mm)", title = "")+
  scale_fill_manual(values = c(rain,sleet,"#b3b3b3"))+
  guides(fill = guide_legend(title = ""))+
  theme_minimal()+
      theme(  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8),
  axis.text.x  = element_text(size = 7),
  axis.text.y  = element_text(size = 7))

```

#### Yearly mean

```{r}
#| label: model-annual-mean-precip
#| echo: false

model_precip_mean_annual <- lm(yearly_precip ~ year, data = data_precip_mean_annual)
summary(model_precip_mean_annual)

data_precip_sum$pred <- predict(model_precip_mean_annual)

```

```{r}
#| label: fig-precipitation-annual
#| echo: false
#| fig-scap: "Mean yearly precipitation (mm)"
#| fig-cap: !expr glue("Mean yearly precipitation (mm) from {min(data_precip_sum$year)} to {max(data_precip_sum$year)}. Data from 2007 is excluded because it was only from May - December. Blue dashed line indicate over all mean of {precip_year_mean} mm. Dashed line represent model prediction.")
#| message: false
#| warning: false

ggplot(data_precip_sum, aes(y = pre_mm_sum, x = year))+
    geom_line(
    aes(y = pred),
    color = "#076834",
    linetype = "dashed",
  ) +
  geom_bar(stat = "identity", fill = "#6e6bc4")+
  scale_x_continuous(breaks=seq(min(data_precip_sum$year), max(data_precip_sum$year), by = 1))+
geom_hline(yintercept = precip_year_mean, linetype="dotted", color = "blue", linewidth =0.5)+
  geom_smooth(method = lm, color = "blue", size = 0.5, se = F, linetype="dashed")+
  labs(x = "", y = "Precipitation (mm)")+
theme_minimal()+
    theme(  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8),
  axis.text.x  = element_text(size = 7),
  axis.text.y  = element_text(size = 7))

```

#### Five year intervals

```{r}
#| label: data-precip-fiveyear
#| echo: false
#| message: false
#| warning: false

data_precip_fiveyear <- data_precip |> 
  filter(year > 2007) |> 
  mutate(
    # choose the anchor year so that 2007–2022 is a bin
    anchor    = 2007,
    seg_start = anchor + floor((year - anchor) / 5) * 5,
    seg_end   = seg_start + 4,
    segment   = paste0(seg_start, "-", seg_end)
  ) |>
  group_by(segment, seg_start, seg_end) |>
  mutate(seg_mean_precip = sum(pre_mm, na.rm = TRUE))

summary(data_precip_fiveyear)

```

# Historic climate data

## Temperature

Data available from Cappelen, J. (2021). *Greenland – DMI Historical Climate Data Collection 1784-2020* (Nos. 21–04; DMI Report). Danish Meteorological Institute. <https://www.dmi.dk/fileadmin/Rapporter/2021/DMIRep21-04.pdf>

Station 4250 is Nuuk, and element 101 is precipitation.

```{r}
#| label: data-import-dmi-historic-data
#| echo: false
#| message: false
#| warning: false

#station 4250 er Nuuk

gr_monthly_all_1784_2020 <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/dmi_historic/DMIRep21-04/gr_monthly_all_1784_2020.csv", delim = ";", escape_double = FALSE, col_types = cols(jan = col_double(), feb = col_double(), mar = col_double(), apr = col_double(), may = col_double(), jun = col_double(), jul = col_double(), aug = col_double(), sep = col_double(), oct = col_double(), nov = col_double(), dec = col_double(), annual = col_double()), locale = locale(decimal_mark = ",", grouping_mark = ""), trim_ws = TRUE)

nuuk_21_04_dmi_temp <- gr_monthly_all_1784_2020 |>
  dplyr::filter(stat_no == 4250, annual != "null", elem_no == 101) |>
  dplyr::mutate(
    dplyr::across(c(jan, feb, mar, apr, may, jun, jul, aug, sep, oct, nov, dec, annual), as.double)) |> 
  drop_na() |> 
  group_by(year) |> 
   mutate(mean_month = mean(c_across(jan:dec)),
          diff = annual - mean_month)

# Define the formula as a function
calculate_temp <- function(year) {
  intercept <- -19.764947
  slope <- 0.009443
  return(intercept + slope * year)
}


annual1784 <- calculate_temp(1784)

annual2024 <- calculate_temp(2024)


nuuk25_08 <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/dmi_historic/DMIRep25-08_1953_2024_data1/425000.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE) |> 
  clean_names() |> 
  dplyr::select(station,year,month,day,hour_utc,x101) |> 
  filter(!is.na(x101)) |> 
  group_by(year) |> 
  summarise(mean_temp = mean(x101))

nuuk21_04_dmi_temp_smoothed <- 
  nuuk_21_04_dmi_temp |>
  arrange(year) |>
  mutate(
    annual_5yr = rollmean(annual, k = 15, fill = NA, align = "center")
  )

nuuk21_04_dmi_temp_smoothed_15yr <- nuuk_21_04_dmi_temp |>
  mutate(
    # choose the anchor year so that 2007–2022 is a bin
    anchor    = 2007,
    seg_start = anchor + floor((year - anchor) / 15) * 15,
    seg_end   = seg_start + 14,
    segment   = paste0(seg_start, "-", seg_end)
  ) |>
  group_by(segment, seg_start, seg_end) |>
  mutate(seg_mean_temp = mean(annual, na.rm = TRUE))

```

```{r}
#| label: model-annual-temp-historic
#| echo: true
#| message: false
#| warning: false

model_annual_temp_historic <- lm(annual ~ year, data = nuuk_21_04_dmi_temp)
summary(model_annual_temp_historic)

nuuk_21_04_dmi_temp$pred <- predict(model_annual_temp_historic)

preds <- predict(model_annual_temp_historic, se.fit = TRUE)

# 2. Create a new dataframe for the ribbon
# We calculate the Upper and Lower bounds (Fit ± 1.96 * SE for 95% CI)
nuuk_21_04_dmi_temp$fit <- preds$fit
nuuk_21_04_dmi_temp$se <- preds$se.fit
nuuk_21_04_dmi_temp$lower <- preds$fit - 1.96 * preds$se.fit
nuuk_21_04_dmi_temp$upper <- preds$fit + 1.96 * preds$se.fit

```

```{r}
#| label: fig-annual-temp-historic-raw-range-filter
#| fig-scap: "Annual historic temperature raw data"
#| fig-cap: !expr glue("Mean air temperature (°C) pr year, from {min(nuuk_21_04_dmi_temp$year)} to {max(nuuk_21_04_dmi_temp$year)}. Yearly mean of {round(mean(nuuk_21_04_dmi_temp$annual, na.rm= TRUE),1)}°C. Yearly mean ranges from {round(min(nuuk_21_04_dmi_temp$annual, na.rm = TRUE), 2)}°C to {max(nuuk_21_04_dmi_temp$annual, na.rm = TRUE)}°C. Modelled mean temperature has increase from {round(annual1784,2)}°C in 1784 to {round(annual2024,2)}°C in 2024.")
#| message: false
#| warning: false

ggplot(data = nuuk_21_04_dmi_temp, aes(x = year, y = annual)) +
    geom_line(
    aes(y = pred),
    color = "#076834",
    linetype = "solid",
  ) +
    geom_ribbon(aes(ymin = lower, ymax = upper), 
              fill = "#076834", 
              alpha = 0.2) + 
  geom_point(size = 3, alpha = 0.7, color = "#076834") +
  #geom_smooth(method = "lm")+
  scale_x_continuous(
    breaks = seq(min(gr_monthly_all_1784_2020$year),
                 max(2025), by = 15))+
  theme_minimal()+
    theme(  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8),
  axis.text.x  = element_text(size = 7),
  axis.text.y  = element_text(size = 7))
```

# Statistical analysis

## Diversity

Data is available from GEM. (2020). *BioBasis Nuuk—Vegetation—NERO line Raunkiaer* (Version 1.0) \[Dataset\]. Greenland Ecosystem Monitoring. <https://doi.org/10.17897/HXJY-RQ13>.

### Importing data and calculating indexes

```{r}
#| label: data-import-indexes
#| echo: true
#| message: false
#| warning: false

#reading the raw datafile
data_raw <- readRDS("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/nero_analysis/data/data_long.rds")

data_filtered <- data_raw |> 
  group_by(year, subsection) |> 
  mutate(no_plots = n_distinct(plot_id)) |> 
  ungroup() |> 
  filter(taxon_code != "rock", no_plots > 1, veg_type != "saltmarsh") |>
  dplyr::select(year,vt_section,subsection, plot_id, no_plots, veg_type, taxon_code, presence)

data_pivotwide <- data_filtered |>
  tidyr::pivot_wider(
    id_cols = c(year, vt_section, subsection, plot_id, no_plots, veg_type),
    names_from = taxon_code,
    values_from = presence,
    values_fill = 0,
    values_fn = max
  )

data_pivotlong <- data_pivotwide |> 
  pivot_longer(cols = carbig:viospe, names_to = "taxon_code", values_to = "presence")

#calculation count and frequency of occurrence
subsection_species <- data_pivotlong |> 
  group_by(year,subsection,veg_type, no_plots, taxon_code) |> 
  summarise(count = sum(presence), foo = count/no_plots)
```

```{r}
#| label: data-calculate-indexes
#| echo: true
#| message: false
#| warning: false

# caluculating richness, evenness and diversity
subsection_indices <- subsection_species |> 
  dplyr::group_by(year, subsection, veg_type) |> 
  dplyr::summarise(
    # restrict to taxa that are present
    H = vegan::diversity(count[count > 0], index = "shannon"),
    S = dplyr::n_distinct(taxon_code[count > 0]),
    shannon = H,
    evenness = ifelse(S > 1, H / log(S), NA_real_),
    richness = S,
    .groups = "drop"
  )

years <- c("2007", "2012", "2017", "2022")
```

Data has been filtered to remove sections with only one remaining plot and subsections in saltmarsh, due to too few replicates in this vegetation type. Final calculates included `r length(unique(subsection_indices$subsection))` subsections (original dataset includes `r length(unique(data_raw$subsection))`).

### Richness

```{r}
#| label: indices-model-richness
#| echo: true
#| message: false
#| warning: false

model_richness_sub_f <- lmer(
  richness ~ factor(year) + (1 | subsection),
  data = subsection_indices
)

summary(model_richness_sub_f)

richness_emmeans_f <- emmeans(model_richness_sub_f, ~ year) 

richness_pairs_f <- emmeans(model_richness_sub_f, ~ year) |> 
  pairs()

richness_emmeans_f
richness_pairs_f

richness_pairs_df <- as.data.frame(richness_pairs_f)
```

```{r}
#| label: fig-indices-model-richness
#| fig-scap: "Species richness"
#| fig-cap: !expr glue("Species richness, number of species pr. subsection. Shared letters indicate non-significant differences among estimated means according to compact letter display.")
#| echo: false
#| message: false
#| warning: false

subsection_indices$year <- factor(subsection_indices$year, levels = years)
richness_emm_df <- as.data.frame(richness_emmeans_f)
richness_emm_df$year <- factor(richness_emm_df$year, levels = years)

richness_cld <- cld(
  richness_emmeans_f,
  adjust = "sidak",
  Letters = letters
)

richness_cld <- as.data.frame(richness_cld)
richness_cld$year <- factor(richness_cld$year, levels = years)

cl_cols <- intersect(c("asymp.LCL", "lower.CL"), names(richness_emm_df))[1]
cu_cols <- intersect(c("asymp.UCL", "upper.CL"), names(richness_emm_df))[1]

ggplot() +
  geom_jitter(
    data = subsection_indices,
    aes(x = year, y = richness),
    width = 0.15,
    alpha = 0.5,
    color = "darkgray"
  ) +
  geom_errorbar(
    data = richness_emm_df,
    aes(x = year, ymin = .data[[cl_cols]], ymax = .data[[cu_cols]]),
    width = 0.25,
    linewidth = 0.9,
    color = "black",
    inherit.aes = FALSE
  ) +
  geom_point(
    data = richness_emm_df,
    aes(x = year, y = emmean),
    size = 2.8,
    color = "black",
    inherit.aes = FALSE
  ) +
  geom_text(
    data = richness_cld,
    aes(x = year, y = max(subsection_indices$richness), label = .group),
    size = 4,
    color = "black",
    inherit.aes = FALSE
  ) +
  labs(x = "", y = "Species richness") +
  theme_minimal()
```

### Evenness

```{r}
#| label: indices-model-evenness
#| echo: true
#| message: false
#| warning: false

model_evenness_sub_f <- lmer(
  evenness ~ factor(year) + (1 | subsection),
  data = subsection_indices
)

summary(model_evenness_sub_f)

evenness_emmeans_f <- emmeans(model_evenness_sub_f, ~ year) 

evenness_pairs_f <- emmeans(model_evenness_sub_f, ~ year) |> 
  pairs()

evenness_emmeans_f
evenness_pairs_f

evenness_pairs_df <- as.data.frame(evenness_pairs_f)
```

```{r}
#| label: fig-indices-model-evenness
#| fig-scap: "Evenness"
#| fig-cap: !expr glue("Evenness pr. subsection. Shared letters indicate non-significant differences among estimated means according to compact letter display.")
#| echo: false
#| message: false
#| warning: false


subsection_indices$year <- factor(subsection_indices$year, levels = years)
evenness_emm_df <- as.data.frame(evenness_emmeans_f)
evenness_emm_df$year <- factor(evenness_emm_df$year, levels = years)

evenness_cld <- cld(
  evenness_emmeans_f,
  adjust = "sidak",
  Letters = letters
)

evenness_cld <- as.data.frame(evenness_cld)
evenness_cld$year <- factor(evenness_cld$year, levels = years)

cl_cols <- intersect(c("asymp.LCL", "lower.CL"), names(evenness_emm_df))[1]
cu_cols <- intersect(c("asymp.UCL", "upper.CL"), names(evenness_emm_df))[1]

ggplot() +
  geom_jitter(
    data = subsection_indices,
    aes(x = year, y = evenness),
    width = 0.15,
    alpha = 0.5,
    color = "darkgray"
  ) +
  geom_errorbar(
    data = evenness_emm_df,
    aes(x = year, ymin = .data[[cl_cols]], ymax = .data[[cu_cols]]),
    width = 0.25,
    linewidth = 0.9,
    color = "black",
    inherit.aes = FALSE
  ) +
  geom_point(
    data = evenness_emm_df,
    aes(x = year, y = emmean),
    size = 2.8,
    color = "black",
    inherit.aes = FALSE
  ) +
  geom_text(
    data = evenness_cld,
    aes(x = year, y = max(subsection_indices$evenness), label = .group),
    size = 4,
    color = "black",
    inherit.aes = FALSE
  ) +
  labs(x = "", y = "Evenness") +
  theme_minimal()
```

### Shannon

### Turnover

## Community composition

### Within vegetation composition

### Between vegetation composition

## Shrubification

### Functional groups

### Specific species
