---
title: "Supplementary Information"
subtitle: "Contains supplementary background figures and tables for figures included in Jacobsen et al . Document is generated from Quato R script enabling recreating of test and figures. Script is available at XXX."
bibliography: references.bib
# author:
#   - name: Ida Bomholt Dyrholm Jacobsen
#     orcid: 0000-0003-2481-8061
#     affiliation: [1, 2]
#     email: ida@example.com
#     corresponding: true
# 
#   - name: Katrine Raundrup
#     orcid: 0000-0002-2110-3709
#     affiliation: 1
# 
#   - name: Niels Martin Schmidt
#     orcid: 0000-0002-4166-6218 
#     affiliation: 2
# 
# affiliations:
#   - id: 1
#     name: Greenland Institute of Natural Resources
#   - id: 2
#     name: Arctic Research Centre, University Y

output:
  pdf_document:
    toc: true
    fig_width: 6.5
    fig_caption: true
    keep_tex: true
    includes:
      in_header: header.tex

format:
  pdf:
    number-sections: true
    number-depth: 2
    pdf-engine: xelatex
    sansfont: "Myriad Pro"
    link-citations: true
    fig-width: 6.5  # Approx 16.5cm, slightly wider than text to be safe
    out-width: 100%
    fig-align: center
    toc: true
    lof: true
    fig-pos: "H"
    include-in-header:
      text: |
        \usepackage{authblk}
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \fancyhf{}
        \lhead{Supplementary Information}
        \rhead{\thepage}
        \renewcommand{\headrulewidth}{0.4pt}
        \setlength{\headheight}{15pt}
        \fancypagestyle{plain}{
          \fancyhf{}
          \lhead{Appendix: NERO line paper}
          \rhead{\thepage}
          \renewcommand{\headrulewidth}{0.4pt}
        }
editor: visual
---

{{< pagebreak >}}

# Current climate conditions 

Data is available 2007 – 2022. 2008 is the first whole year of temperature data. Data available from https://data.g-e-m.dk [@gem2026d; @gem2026c].

## Temperature

```{r}
#| label: importing-packages
#| output: false
#| echo: false
#| message: false
#| warning: false

#install.packages("rmarkdown","pacman","tidyverse","yaml", dependencies = TRUE)
library(pacman)
pacman::p_load(ggplot2, Hmisc, tidyverse, janitor, lubridate, glue, hms, knitr, broom, ggpmisc, yaml, readxl, zoo, emmeans, lme4, multcomp, vegan, lmerTes, codyn, patchwork, tibble, broom.mixed, ggeffects, systemfonts, ragg, pbkrtest)

# pbkrtest dependincy for the species mode

opts_chunk$set(size = "small", dev = "cairo_pdf")

# colors
sleet <- rgb(110/255, 107/255, 196/255, 1)
rain <- rgb(0/255, 0/255, 165/255, 1)
warm <- rgb(165/255, 0/255, 0/255, 1)
fig_text_size = 11

# global font settings
theme_set(
  theme_minimal(base_family = "Myriad Pro", base_size = 8) +
    theme(  
          axis.title.x = element_text(size = 10),
          axis.title.y = element_text(size = 10),
          axis.text.x  = element_text(size = 8),
          axis.text.y  = element_text(size = 8))
)

cld_table <- function(cld_df, caption) {
  cld_df <- as.data.frame(cld_df)
  cl <- intersect(c("lower.CL", "asymp.LCL"), names(cld_df))[1]
  cu <- intersect(c("upper.CL", "asymp.UCL"), names(cld_df))[1]
  cld_df |>
    dplyr::mutate(group = trimws(.group)) |>
    dplyr::select(year, emmean, SE,
                  lower = dplyr::all_of(cl),
                  upper = dplyr::all_of(cu),
                  group) |>
    dplyr::arrange(year) |>
    knitr::kable(digits = 3, caption = caption)
}
```

```{r}
#| label: importing-temp-data
#| echo: false
#| output: false
#| message: false
#| warning: false

data_temp <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/temperature/data/View_ClimateBasis_Nuuk_Data_Temperature_Air_temperature_@_200_cm__30min_average__DegreesC220420240157436123.csv", delim = "\t", escape_double = FALSE,col_types = cols(Date = col_datetime(format = "%Y-%m-%d"), Time = col_time(format = "%H:%M:%S")), trim_ws = TRUE) |> 
  janitor::clean_names() |> 
  drop_na() |>  
  filter(at_c != -9999) |>  
   mutate(year = year(date),
         month = month(date), 
         doy = yday(date),
         month_string = month.name[month])

temp_year_mean <- round(mean(data_temp$at_c),4)

data_temp_monthly_mean <- data_temp |> 
  group_by(month) |> 
  dplyr::reframe(mean = round(mean(at_c),2),
            sd = round(sd(at_c),2)) |> 
  mutate(month_string = month.name[month]) |> 
  arrange(desc(mean)) |> 
  mutate(temp_rank = row_number())

data_temp_month_mean <- data_temp_monthly_mean

```

```{r}
#| label: temp-mean-monthly
#| echo: false
#| output: false
#| message: false
#| warning: false


data_temp_annual_mean <- data_temp |> 
  group_by(year) |> 
  filter(year != 2007) |> #excluding 2007 because it is only the months of oct, nov, dec
  dplyr::reframe(mean = mean(at_c),
            sd = sd(at_c))

```

#### Monthly mean temperature

```{r}
#| label: fig-month-mean-temp
#| fig-scap: "Mean air temperature (°C) per month (Background)"
#| fig-cap: !expr glue("Mean temperature per month. Mean air temperature (°C) per month, based on data from {min(data_temp$year)} to {max(data_temp$year)}. The dashed line represent yearly mean of {round(temp_year_mean,3)}°C.")
#| echo: false
#| output: true
#| message: false
#| warning: false

ggplot(data_temp_month_mean, aes(x = month, y = mean))+
  geom_line(linewidth = 0.25, linetype="solid", alpha = 0.2)+
  geom_errorbar(
    aes(ymin = mean - sd, ymax = mean + sd),
    width = 0.2, linewidth = 0.25, alpha = 0.4
  ) +
  geom_smooth(method = "loess", formula = y ~ x, color = "gray", se = FALSE, alpha = 0.1)+
  geom_point(size = 2, alpha = 0.7, color = "#076834") +
    scale_x_continuous(breaks=seq(1, 12, by = 1))+
  geom_hline(yintercept=temp_year_mean, linetype="dotted", color = "#076834", linewidth = 0.5)+
  labs(x = "Month", y = "Mean temperature (°C)", title = "")
```

The mean annual temperature is `r round(temp_year_mean, 1)` °C.
The warmest months are `r data_temp_month_mean$month_string[1]` (`r round(data_temp_month_mean$mean[1], 1)` °C), `r data_temp_month_mean$month_string[2]` (`r round(data_temp_month_mean$mean[2], 1)` °C), `r data_temp_month_mean$month_string[3]` (`r round(data_temp_month_mean$mean[3], 1)` °C).
The coldest months are `r data_temp_month_mean$month_string[10]` (`r round(data_temp_month_mean$mean[10], 1)` °C), `r data_temp_month_mean$month_string[11]` (`r round(data_temp_month_mean$mean[11], 1)` °C), `r data_temp_month_mean$month_string[12]` (`r round(data_temp_month_mean$mean[12], 1)` °C).

{{< pagebreak >}}

#### Yearly mean temperature

```{r}
#| label: model-annual-mean-temp
#| echo: false
#| output: false
#| message: false
#| warning: false

model_temp_annual_mean <- lm(mean ~ year, data = data_temp_annual_mean)
summary(model_temp_annual_mean)

data_temp_annual_mean$pred <- predict(model_temp_annual_mean)

```

```{r}
#| label: fig-annual-mean-temp
#| fig-scap: "Mean air temperature per year (°C) (Background)"
#| fig-cap: !expr glue("Mean annual air temperature per year (°C) from {min(data_temp_annual_mean$year)} to {max(data_temp_annual_mean$year)} in Kangerluassunguaq. Data from 2007 is excluded because data is only from October, November and December. Gray dotted line indicate over all mean of {temp_year_mean}°C. Dashed line is model prediction.")
#| echo: false
#| output: true
#| message: false
#| warning: false

ggplot(data_temp_annual_mean, aes(x = year, y = mean))+
  geom_line(linewidth = 0.4, color = "gray", linetype="solid", alpha = 0.5)+
  #geom_smooth(method = lm, color = "#076834", se = F, linetype = "dashed")+
  geom_line(
    aes(y = pred),
    color = "#076834",
    linetype = "dashed"
  ) +
  geom_point(size = 2, alpha = 0.7, color = "#076834") +
    scale_x_continuous(breaks=seq(min(data_temp_annual_mean$year), max(data_temp_annual_mean$year), by = 1))+
  geom_hline(yintercept=temp_year_mean, linetype="dotted", color = "darkgray", linewidth=0.7)+
  labs(x = "", y = "Mean temperature (°C)", title = "")
```

{{< pagebreak >}}

#### Five year intervals of temperature

```{r}
#| label: data-fiveyear_intervals
#| echo: false
#| output: false
#| message: false
#| warning: false

data_temp_fiveyear <- data_temp |> 
  mutate(
    # choose the anchor year so that 2007–2022 is a bin
    anchor    = 2007,
    seg_start = anchor + floor((year - anchor) / 5) * 5,
    seg_end   = seg_start + 4,
    segment   = paste0(seg_start, "-", seg_end)
  ) |>
  group_by(segment, seg_start, seg_end) |>
  mutate(seg_mean_temp = mean(at_c, na.rm = TRUE))

interval_lines <- data_temp_fiveyear |>
  filter(year > 2007) |> 
  distinct(segment, seg_start, seg_end, seg_mean_temp)

```

```{r}
#| label: data-fiveyear_intervals_test
#| echo: false
#| output: false
#| message: false
#| warning: false

# Fit the linear mixed model
temp_data <- data_temp_fiveyear |> filter(year > 2007)

# Fit the model explicitly
model_temp_fiveyear <- lmer(
  at_c ~ segment + (1 | year), 
  data = temp_data
)

# Get estimated marginal means with cld
emm_temp_fiveyear <- model_temp_fiveyear |>
  emmeans(~ segment) |>
  multcomp::cld(alpha = 0.05, Letters = letters, adjust = "tukey")

cld_temp <- as.data.frame(emm_temp_fiveyear)

cld_plot <- cld_temp |>
  left_join(interval_lines, by = "segment")

```

```{r}
#| label: fig-temp-fiveyear
#| fig-scap: "5 year intervals in temperature means (Background)"
#| fig-cap: !expr glue("Five year intervals in temperature means. Shared letters indicate non-significant differences among estimated marginal means according to a compact letter display .")
#| echo: false
#| output: true
#| message: false
#| warning: false

yearly_means <- data_temp_fiveyear |>
  group_by(year) |>
  filter(year > 2007) |> 
  summarise(
    mean_temp = mean(at_c, na.rm = TRUE),
    .groups = "drop"
  )

interval_lines <- data_temp_fiveyear |>
  filter(year > 2007) |> 
  distinct(segment, seg_start, seg_end, seg_mean_temp)

# amke the positions of the cld in the plot
cld_plot <- cld_plot |>
  mutate(
    x_mid = (seg_start + seg_end) / 2
  )

# 3. Plot
ggplot() +

  # Reference line: yearly grand mean
  geom_hline(
    yintercept = temp_year_mean,
    linetype   = "dotted",
    color      = "darkgray",
    linewidth  = 0.7
  ) +

  # Base layer: yearly means
  geom_point(
    data  = yearly_means,
    aes(x = year, y = mean_temp),
    size  = 2,
    color = "darkgreen"
  ) +

  # Interval mean lines
  geom_segment(
    data = interval_lines,
    aes(
      x    = seg_start,
      xend = seg_end,
      y    = seg_mean_temp,
      yend = seg_mean_temp
    )
  ) +

  # Compact letter display (CLD)
  geom_text(
    data = cld_plot,
    aes(
      x     = x_mid,
      y     = max(yearly_means$mean_temp),
      label = .group
    ),
    vjust = -0.8,
    size  = 3,
    nudge_x = 0.2
  ) +

  # Scales
  scale_x_continuous(
    breaks = seq(min(yearly_means$year), max(yearly_means$year), by = 1)
  ) +

  # Labels
  labs(
    x     = "",
    y     = "Mean Annual Temperature (°C)",
    title = ""
  )
```

{{< pagebreak >}}

## Precipitation

Data available from https://data.g-e-m.dk [@gem2026c]. 

```{r}
#| label: importing-precip
#| echo: false
#| output: false
#| message: false
#| warning: false

data_precip <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/precipitation/Precipitation accumulated (mm)_10.17897_SXJ8-WA79 2/Precipitation accumulated (mm)_10.17897_SXJ8-WA79_data.txt", delim = "\t", escape_double = FALSE, col_types = cols(Date = col_date(format = "%Y-%m-%d"), Time = col_time(format = "%H:%M:%S")), trim_ws = TRUE) |> 
  janitor::clean_names() |> 
  drop_na() |>
  filter(pre_mm != -9999) |> 
   mutate(year = year(date),
         month = month(date),
         doy = yday(date),
         month_string = month.name[month])
```

```{r}
#| label: data-precip-mean-calculations
#| echo: false
#| output: false
#| message: false
#| warning: false

data_precip_sum <- data_precip |> 
  group_by(year) |> 
  filter(year != 2007) |> 
  summarise(pre_mm_sum = sum(pre_mm))

data_precip_mean <- data_precip_sum |> 
  summarize(mean = mean(pre_mm_sum))

precip_year_mean <- round(mean(data_precip_mean$mean),2)

data_precip_mean_year <- data_precip  |> 
  group_by(year) |> 
  summarise(yearly_precip = sum(pre_mm, na.rm = TRUE)) |> 
  summarise(mean_precip = mean(yearly_precip, na.rm = TRUE))  |> 
  pull(mean_precip)

data_temp_hour <- data_temp |> 
  mutate(hour = substr(time, 1, 2)) |> 
  group_by(date, hour) |> 
  summarise(at_c = mean(at_c, na.rm = TRUE)) |> 
  mutate(time = hms::hms(hours = as.numeric(hour)))

sleet_lower <- -1
sleet_upper <- 1 
  
  data_pre_tem_joined <- data_precip |> 
  left_join(data_temp_hour, by = c("date" = "date", "time" = "time")) |> 
  drop_na() |> 
  mutate(precip_type = case_when(at_c < sleet_lower ~ "snow",
                              at_c <= sleet_upper ~ "sleet",
                              TRUE ~ "rain"))
  
  data_precip_month_sum <- data_precip |> 
  group_by(month, year) |> 
  summarise(pre_mm_sum = sum(pre_mm))

data_precip_month_mean <- data_precip_month_sum |> 
  group_by(month) |> 
  summarise(pre_mm_mean = round(mean(pre_mm_sum),2),
         sd = sd(pre_mm_mean)) |> 
  mutate(month_string = month.name[month]) |> 
  arrange(desc(pre_mm_mean)) |> 
  mutate(temp_rank = row_number())

precip_month_mean <- round(mean(data_precip_month_mean$pre_mm_mean),2)

data_precip_mean_annual <- data_precip  |> 
  filter(year != 2007) |> 
  group_by(year) |> 
  summarise(yearly_precip = sum(pre_mm, na.rm = TRUE))

precip_type_pct <- data_pre_tem_joined |>
  filter(year > 2007) |>
  group_by(month, precip_type) |>
  summarise(total = sum(pre_mm, na.rm = TRUE), .groups = "drop") |>
  group_by(month) |>
  mutate(pct = total / sum(total) * 100)

```

The mean annual precipitation is `r precip_year_mean` mm.

The months with the most precipitation are `r data_precip_month_mean$month_string[1]` (`r data_precip_month_mean$pre_mm_mean[1]` mm), `r data_precip_month_mean$month_string[2]` (`r data_precip_month_mean$pre_mm_mean[2]` mm), `r data_precip_month_mean$month_string[3]` (`r data_precip_month_mean$pre_mm_mean[3]` mm), which account for `r data_precip_month_mean$pre_mm_mean[1]+data_precip_month_mean$pre_mm_mean[2]+data_precip_month_mean$pre_mm_mean[3]` mm or `r  round((data_precip_month_mean$pre_mm_mean[1]+data_precip_month_mean$pre_mm_mean[2]+data_precip_month_mean$pre_mm_mean[3])/precip_year_mean*100,1)` % of the yearly precipitation.

The months with the least precipitation are `r data_precip_month_mean$month_string[10]` (`r data_precip_month_mean$pre_mm_mean[10]` mm), `r data_precip_month_mean$month_string[11]` (`r data_precip_month_mean$pre_mm_mean[11]` mm), `r data_precip_month_mean$month_string[12]` (`r data_precip_month_mean$pre_mm_mean[12]` mm).

#### Monthly mean precipitation

```{r}
#| label: data-yearly-mean-dist_rain-snow
#| echo: false
#| output: false
#| message: false
#| warning: false

# joint the temperature precipitation data to distinguish snow and rain. 

# Total precipitation per month from full dataset
data_precip_monthly <- data_precip |>
  filter(year != 2007) |>
  group_by(year, month) |>
  summarise(sum = sum(pre_mm, na.rm = TRUE), .groups = "drop") |>
  group_by(month) |>
  summarise(mean = round(mean(sum), 2), .groups = "drop")

# % distribution of precip type per month from joined dataset
precip_type_pct <- data_pre_tem_joined |>
  filter(year != 2007) |>
  group_by(month, precip_type) |>
  summarise(total = sum(pre_mm, na.rm = TRUE), .groups = "drop") |>
  group_by(month) |>
  mutate(pct = total / sum(total) * 100)
```

```{r}
#| label: fig-snow-rain-year
#| fig-scap: "Mean monthly distribution of types of precipitation (mm) (Background)"
#| fig-cap: !expr glue("Mean precipitation type per month. Mean monthly distribution of types of precipitation (mm). Totals based on data from {min(data_precip$year)} to {max(data_precip$year)} (requires join with temperature data which it only available till {max(precip_type_pct$year)}). Sleet is defined as precipitation that fell when temperatures were between {sleet_lower}°C and {sleet_upper}°C.")
#| echo: false
#| output: true
#| message: false
#| warning: false

# Scale the type percentages against the full monthly totals
precip_type_scaled <- precip_type_pct |>
  left_join(data_precip_monthly, by = "month") |>
  mutate(scaled_mean = pct / 100 * mean)

# Plot
ggplot(precip_type_scaled, aes(x = month, y = scaled_mean, fill = precip_type)) +
  geom_bar(
    stat  = "identity"
  ) +
  scale_x_continuous(
    breaks = seq(1, 12, by = 1)
  ) +
  scale_fill_manual(values = c(rain, sleet, "#b3b3b3")) +
  guides(fill = guide_legend(title = "")) +
  labs(
    x     = "",
    y     = "Mean precipitation (mm)",
    title = ""
  )

```

{{< pagebreak >}}

#### Yearly mean precipitation

```{r}
#| label: model-annual-mean-precip
#| echo: false
#| output: false
#| message: false
#| warning: false

model_precip_mean_annual <- lm(yearly_precip ~ year, data = data_precip_mean_annual)
summary(model_precip_mean_annual)

data_precip_sum$pred <- predict(model_precip_mean_annual)

```

```{r}
#| label: fig-precipitation-annual
#| fig-scap: "Mean yearly precipitation (mm) (Background)"
#| fig-cap: !expr glue("Mean yearly precipitation (mm) from {min(data_precip_sum$year)} to {max(data_precip_sum$year)}. Data from 2007 is excluded because it was only from May - December. Blue dashed line indicate over all mean of {precip_year_mean} mm. Dashed line represent model prediction.")
#| echo: false
#| output: true
#| message: false
#| warning: false

ggplot(data_precip_sum, aes(y = pre_mm_sum, x = year))+
  geom_hline(yintercept = precip_year_mean, linetype="dotted", color = "darkgray", linewidth =0.5)+
  geom_bar(stat = "identity", fill = "#6e6bc4")+
  scale_x_continuous(breaks=seq(min(data_precip_sum$year), max(data_precip_sum$year), by = 1))+
  geom_line(
      aes(y = pred),
      color = "#076834",
      linetype = "dashed")+
  labs(x = "", y = "Precipitation (mm)")

#geom_smooth(method = lm, color = "blue", size = 0.5, se = F, linetype="dashed")+
#  

```

#### Five year interval precipitation

```{r}
#| label: data-precip-fiveyear
#| echo: false
#| output: false
#| message: false
#| warning: false

data_precip_fiveyear <- data_precip |> 
  filter(year > 2007) |> 
  mutate(
    # choose the anchor year so that 2007–2022 is a bin
    anchor    = 2007,
    seg_start = anchor + floor((year - anchor) / 5) * 5,
    seg_end   = seg_start + 4,
    segment   = paste0(seg_start, "-", seg_end)
  ) |>
  group_by(segment, seg_start, seg_end) |>
  mutate(seg_mean_precip = sum(pre_mm, na.rm = TRUE))


# Annual totals from hourly data
yearly_precip <- data_precip_fiveyear |>
  group_by(year, segment, seg_start, seg_end) |>
  summarise(
    annual_precip = sum(pre_mm, na.rm = TRUE),
    .groups = "drop"
  )

# Segment totals (sum of annual totals per interval)
interval_lines_precip <- yearly_precip |>
  group_by(segment, seg_start, seg_end) |>
  summarise(
    seg_total_precip = mean(annual_precip, na.rm = TRUE),
    .groups = "drop"
  )

# Grand mean annual precipitation (reference line)
precip_year_mean <- mean(yearly_precip$annual_precip, na.rm = TRUE)

# CLD positions
# cld_plot_precip <- cld_plot_precip |>
#   mutate(
#     x_mid = (seg_start + seg_end) / 2
#   )

```

```{r}
#| label: precip-fiveyear-intervals-test
#| echo: false
#| output: false
#| message: false
#| warning: false

#The residuals are not well-approximated by a normal distribution. The Shapiro-Wilk p-value is borderline but the Q-Q plot shows meaningful tail deviation, and with this sample size the test is underpowered to detect it.

kruskal.test(annual_precip ~ segment, data = yearly_precip)

```

```{r}
#| label: fig-precip-fiveyear
#| fig-scap: "5 year intervals of precipitation means (Background)"
#| fig-cap: !expr glue("Five year intervals in precipitation means. Shared letters indicate non-significant differences among estimated marginal means according to a compact letter display based on Tukey-adjusted pairwise comparisons at the 5% significance level.")
#| echo: false
#| output: true
#| message: false
#| warning: false


# Interval lines from segment summaries
interval_lines_precip <- yearly_precip |>
  group_by(segment, seg_start, seg_end) |>
  summarise(
    seg_total_precip = mean(annual_precip, na.rm = TRUE),
    .groups = "drop"
  )

# Grand mean reference line
precip_year_mean <- mean(yearly_precip$annual_precip, na.rm = TRUE)

# Plot
ggplot() +
  # Reference line: yearly grand mean
  geom_hline(
    yintercept = precip_year_mean,
    linetype   = "dotted",
    color      = "darkgray",
    linewidth  = 0.7
  ) +
  # Base layer: yearly totals
    geom_bar(
    data  = yearly_precip,
    aes(x = year, y = annual_precip),
    stat  = "identity",
    fill  = "#6e6bc4",
    alpha = 0.7
  ) +
  # Interval mean lines
  geom_segment(
    data = interval_lines_precip,
    aes(
      x    = seg_start,
      xend = seg_end,
      y    = seg_total_precip,
      yend = seg_total_precip
    )
  ) +
  # Scales
  scale_x_continuous(
    breaks = seq(min(yearly_precip$year), max(yearly_precip$year), by = 1)
  ) +
  # Labels
  labs(
    x     = "",
    y     = "Annual Precipitation (mm)",
    title = ""
  )
```

{{< pagebreak >}}

## Radiation

Data available from https://data.g-e-m.dk [@gem2026b]

```{r}
#| label: data-import-radiation
#| echo: false
#| output: false
#| message: false
#| warning: false

par <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/ration_par/View_ClimateBasis_Nuuk_Data_Radiation_Photosyntetic_active_radiation_@_200_cm__5min_average_æmol_m2_sec210420241337080578.csv", delim = "\t", escape_double = FALSE, trim_ws = TRUE) |> 
  clean_names() |> 
  filter(par_mmol_sec_m2 != -9999, solar_elevation > 0) |> 
    mutate(
     year = year(date),
     month = month(date)) |> 
      mutate(
    anchor    = 2007,
    seg_start = anchor + floor((year - anchor) / 5) * 5,
    seg_end   = seg_start + 4,
    segment   = paste0(seg_start, "-", seg_end)
  ) |> 
  group_by(segment, seg_start, seg_end) |>
  mutate(seg_mean_par = mean(par_mmol_sec_m2, na.rm = TRUE))

par_year <- par |> 
  group_by(year) |> 
  reframe(year_mean = mean(par_mmol_sec_m2))

par_segment <- par |>   
    mutate(
    anchor    = 2007,
    seg_start = anchor + floor((year - anchor) / 5) * 5,
    seg_end   = seg_start + 4,
    segment   = paste0(seg_start, "-", seg_end)
  ) |> 
  group_by(segment, seg_start, seg_end) |>
  reframe(seg_mean_par = mean(par_mmol_sec_m2, na.rm = TRUE))

par_month <- par |> 
  group_by(month) |> 
  reframe(month_mean = mean(par_mmol_sec_m2))

```

```{r}
#| label: model-fiveyear-par
#| echo: false
#| output: false
#| message: false
#| warning: false

# Fit the model explicitly
model_par_fiveyear <- lmer(
  par_mmol_sec_m2 ~ segment + (1 | year), 
  data = par
)

# Get estimated marginal means with cld
emm_par_fiveyear <- model_par_fiveyear |>
  emmeans(~ segment) |>
  cld(alpha = 0.05, Letters = letters, adjust = "tukey")

cld_par <- as.data.frame(emm_par_fiveyear)

cld_plot_par <- cld_par |>
  left_join(par_segment, by = "segment")

```

```{r}
#| label: fig-par-fiveyear
#| fig-scap: "PAR means, 5 year interval (Background)"
#| fig-cap: !expr glue("Five year intervals in PAR means. Shared letters indicate non-significant differences among estimated marginal means according to a compact letter display based on Tukey-adjusted pairwise comparisons at the 5% significance level.")
#| echo: false
#| output: true
#| message: false
#| warning: false

yearly_means_par <- par |>
  group_by(year) |>
  summarise(
    mean_par = mean(par_mmol_sec_m2, na.rm = TRUE),
    .groups = "drop"
  )

interval_lines_par <- par |>
  distinct(segment, seg_start, seg_end, seg_mean_par)

cld_plot_par <- cld_plot_par |>
  mutate(
    x_mid = (seg_start + seg_end) / 2
  )

overall_mean_par <- mean(par$par_mmol_sec_m2, na.rm = TRUE)

ggplot() +
  geom_hline(yintercept = overall_mean_par, linetype = "dotted", color = "darkgray", linewidth = 0.7) +
  geom_point(
    data = yearly_means_par,
    aes(x = year, y = mean_par),
    size = 2,
    color = "darkgreen"
  ) +
  geom_segment(
    data = interval_lines_par,
    aes(
      x = seg_start,
      xend = seg_end,
      y = seg_mean_par,
      yend = seg_mean_par
    )
  ) +
  geom_text(
    data = cld_plot_par,
    aes(
      x = x_mid,
      y = max(yearly_means_par$mean_par)*0.99,
      label = .group
    ),
    vjust = -0.8,
    size = 3, nudge_x = 0.2
  ) +
  labs(
    x = "",
    y = expression("Mean Annual PAR (" * mu * "mol m"^-2 * " s"^-1 * ")"),
    title = ""
  ) +
  scale_x_continuous(breaks = seq(min(yearly_means_par$year), max(yearly_means_par$year), by = 1))
```

{{< pagebreak >}}

# Historic climate data

## Temperature

Data available from Cappelen [-@cappelen2021].

Station 4250 is Nuuk, and element 101 is precipitation.

```{r}
#| label: data-import-dmi-historic-data
#| echo: false
#| output: false
#| message: false
#| warning: false

#station 4250 er Nuuk

gr_monthly_all_1784_2020 <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/dmi_historic/DMIRep21-04/gr_monthly_all_1784_2020.csv", delim = ";", escape_double = FALSE, col_types = cols(jan = col_double(), feb = col_double(), mar = col_double(), apr = col_double(), may = col_double(), jun = col_double(), jul = col_double(), aug = col_double(), sep = col_double(), oct = col_double(), nov = col_double(), dec = col_double(), annual = col_double()), locale = locale(decimal_mark = ",", grouping_mark = ""), trim_ws = TRUE)

nuuk_21_04_dmi_temp <- gr_monthly_all_1784_2020 |>
  dplyr::filter(stat_no == 4250, annual != "null", elem_no == 101) |>
  dplyr::mutate(
    dplyr::across(c(jan, feb, mar, apr, may, jun, jul, aug, sep, oct, nov, dec, annual), as.double)) |> 
  drop_na() |> 
  group_by(year) |> 
   mutate(mean_month = mean(c_across(jan:dec)),
          diff = annual - mean_month)

# Define the formula as a function
calculate_temp <- function(year) {
  intercept <- -19.764947
  slope <- 0.009443
  return(intercept + slope * year)
}


annual1784 <- calculate_temp(1784)

annual2024 <- calculate_temp(2024)


nuuk25_08 <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/dmi_historic/DMIRep25-08_1953_2024_data1/425000.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE) |> 
  clean_names() |> 
  dplyr::select(station,year,month,day,hour_utc,x101) |> 
  filter(!is.na(x101)) |> 
  group_by(year) |> 
  summarise(mean_temp = mean(x101))

nuuk21_04_dmi_temp_smoothed <- 
  nuuk_21_04_dmi_temp |>
  arrange(year) |>
  mutate(
    annual_5yr = rollmean(annual, k = 15, fill = NA, align = "center")
  )

nuuk21_04_dmi_temp_smoothed_15yr <- nuuk_21_04_dmi_temp |>
  mutate(
    # choose the anchor year so that 2007–2022 is a bin
    anchor    = 2007,
    seg_start = anchor + floor((year - anchor) / 15) * 15,
    seg_end   = seg_start + 14,
    segment   = paste0(seg_start, "-", seg_end)
  ) |>
  group_by(segment, seg_start, seg_end) |>
  mutate(seg_mean_temp = mean(annual, na.rm = TRUE))

```

```{r}
#| label: model-annual-temp-historic
#| echo: false
#| output: false
#| message: false
#| warning: false

model_annual_temp_historic <- lm(annual ~ year, data = nuuk_21_04_dmi_temp)
summary(model_annual_temp_historic)

nuuk_21_04_dmi_temp$pred <- predict(model_annual_temp_historic)

preds <- predict(model_annual_temp_historic, se.fit = TRUE)

# 2. Create a new dataframe for the ribbon
# We calculate the Upper and Lower bounds (Fit ± 1.96 * SE for 95% CI)
nuuk_21_04_dmi_temp$fit <- preds$fit
nuuk_21_04_dmi_temp$se <- preds$se.fit
nuuk_21_04_dmi_temp$lower <- preds$fit - 1.96 * preds$se.fit
nuuk_21_04_dmi_temp$upper <- preds$fit + 1.96 * preds$se.fit

```

```{r}
#| label: fig-annual-temp-historic-raw-range-filter
#| fig-scap: "Annual historic temperature raw data (Background)"
#| fig-cap: !expr glue("Historic yearly mean air temperature (°C) in Nuuk, from {min(nuuk_21_04_dmi_temp$year)} to {max(nuuk_21_04_dmi_temp$year)}. Yearly mean of {round(mean(nuuk_21_04_dmi_temp$annual, na.rm= TRUE),1)}°C. Yearly mean ranges from {round(min(nuuk_21_04_dmi_temp$annual, na.rm = TRUE), 2)}°C to {max(nuuk_21_04_dmi_temp$annual, na.rm = TRUE)}°C. Modelled mean temperature has increase from {round(annual1784,2)}°C in 1784 to {round(annual2024,2)}°C in 2024 (increase of {round(annual2024,2)-round(annual1784,2)}°C or {round((round(annual2024,2)-round(annual1784,2))/((2024-1784)/10),3)}°C/decade.)")
#| echo: false
#| output: true
#| message: false
#| warning: false

ggplot(data = nuuk_21_04_dmi_temp, aes(x = year, y = annual)) +
    geom_line(
    aes(y = pred),
    color = "darkgray",
    linetype = "solid",
  ) +
    geom_ribbon(aes(ymin = lower, ymax = upper), 
              fill = "darkgray", 
              alpha = 0.2) + 
  geom_point(size = 2, alpha = 0.7, color = "darkgreen") +
  #geom_smooth(method = "lm")+
  scale_x_continuous(
    breaks = seq(min(gr_monthly_all_1784_2020$year),
                 max(2025), by = 15))+
  labs(x = "", y = "Temperature (°C)")
```

{{< pagebreak >}}

# Statistical analysis

## Diversity

Data available from https://data.g-e-m.dk [@gem2026e].


```{r}
#| label: data-import-indexes
#| echo: false
#| output: false
#| message: false
#| warning: false

#Importing data and calculating indexes

#reading the raw datafile
data_raw <- readRDS("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/nero_analysis/data/data_long.rds")

data_filtered <- data_raw |> 
  group_by(year, subsection) |> 
  mutate(no_plots = n_distinct(plot_id)) |> 
  ungroup() |> 
  filter(taxon_code != "rock", no_plots > 1, veg_type != "saltmarsh") |>
  dplyr::select(year,vt_section,subsection, plot_id, no_plots, veg_type, taxon_code, presence)

data_pivotwide <- data_filtered |>
  tidyr::pivot_wider(
    id_cols = c(year, vt_section, subsection, plot_id, no_plots, veg_type),
    names_from = taxon_code,
    values_from = presence,
    values_fill = 0,
    values_fn = max
  )

data_pivotlong <- data_pivotwide |> 
  pivot_longer(cols = carbig:viospe, names_to = "taxon_code", values_to = "presence")

#calculation count and frequency of occurrence
subsection_species <- data_pivotlong |> 
  group_by(year,subsection,veg_type, no_plots, taxon_code) |> 
  summarise(count = sum(presence), foo = count/no_plots) |> 
  distinct()

subsection_species_wide <- subsection_species |>
  dplyr::select(-count,-no_plots) |> 
  pivot_wider(names_from = taxon_code, values_from = foo, values_fill = 0)

```

```{r}
#| label: data-calculate-indexes
#| echo: false
#| output: false
#| message: false
#| warning: false

# caluculating richness, evenness and diversity
subsection_indices <- subsection_species |> 
  dplyr::group_by(year, subsection, veg_type) |> 
  dplyr::summarise(
    # restrict to taxa that are present
    H = vegan::diversity(count[count > 0], index = "shannon"),
    S = dplyr::n_distinct(taxon_code[count > 0]),
    shannon = H,
    evenness = ifelse(S > 1, H / log(S), NA_real_),
    richness = S,
    .groups = "drop"
  )

years <- c("2007", "2012", "2017", "2022")
```

Data has been filtered to remove sections with only one remaining plot and subsections in saltmarsh, due to too few replicates in this community type. Final calculates included `r length(unique(subsection_indices$subsection))` subsections (full transect consists of 162 subsection).

### Richness

```{r}
#| label: indices-model-richness
#| echo: false
#| output: false
#| message: false
#| warning: false

model_richness_sub_f <- lmer(
  richness ~ factor(year) + (1 | subsection),
  data = subsection_indices
)

#summary(model_richness_sub_f)

richness_emmeans_f <- emmeans(model_richness_sub_f, ~ year) 

richness_pairs_f <- emmeans(model_richness_sub_f, ~ year) |> 
  pairs()

richness_emmeans_f
richness_pairs_f

richness_pairs_df <- as.data.frame(richness_pairs_f)
```

```{r}
#| label: fig-indices-model-richness
#| fig-scap: "Species richness (Part a, Figure 3)"
#| fig-cap: !expr glue("Species richness, number of species pr. subsection. Shared letters indicate non-significant differences among estimated marginal means according to a compact letter display based on Tukey-adjusted pairwise comparisons at the 5% significance level.")
#| echo: false
#| output: true
#| message: false
#| warning: false

subsection_indices$year <- factor(subsection_indices$year, levels = years)
richness_emm_df <- as.data.frame(richness_emmeans_f)
richness_emm_df$year <- factor(richness_emm_df$year, levels = years)

richness_cld <- cld(
  richness_emmeans_f,
  adjust = "tukey",
  Letters = letters
)

jitter_width = 0.06
crossbar_width = 0.25

richness_cld <- as.data.frame(richness_cld)
richness_cld$year <- factor(richness_cld$year, levels = years)

cl_cols <- intersect(c("asymp.LCL", "lower.CL"), names(richness_emm_df))[1]
cu_cols <- intersect(c("asymp.UCL", "upper.CL"), names(richness_emm_df))[1]

plot_richness <- ggplot() +
  geom_jitter(
    data = subsection_indices,
    aes(x = year, y = richness),
    width = jitter_width,
    alpha = 0.4,
    color = "darkgreen",
    size = 2
  ) +
    geom_crossbar(data = richness_emm_df,
              aes(x = year, y = emmean,
                  ymin = .data[[cl_cols]], 
                  ymax = .data[[cu_cols]]),
              width = crossbar_width,
              fill = "black", 
              alpha = 0.1,      # <<< no fill
              colour = "black",   # <<< outline back
              linewidth = 0.2)  +  # <<< very thin
  # geom_point(data = richness_emm_df,
  #            aes(x = year, y = emmean),
  #            size = 7, colour = "black", shape = 95) +
  geom_text(
    data = richness_cld,
    aes(x = year, y = max(subsection_indices$richness), label = .group),
    size = 3,
    color = "black",
    inherit.aes = FALSE, nudge_x = 0.2
  ) +
  labs(x = "", y = "Species richness") 

plot_richness
```
```{r}
#| label: indices-tbl-richness
#| echo: false
#| output: true
#| message: false
#| warning: false


cld_table(richness_cld,  "Estimated marginal means and CLD for species richness per survey year")

```

### Evenness

```{r}
#| label: indices-model-evenness
#| echo: false
#| output: false
#| message: false
#| warning: false

model_evenness_sub_f <- lmer(
  evenness ~ factor(year) + (1 | subsection),
  data = subsection_indices
)

summary(model_evenness_sub_f)

evenness_emmeans_f <- emmeans(model_evenness_sub_f, ~ year) 

evenness_pairs_f <- emmeans(model_evenness_sub_f, ~ year) |> 
  pairs()

evenness_emmeans_f
evenness_pairs_f

evenness_pairs_df <- as.data.frame(evenness_pairs_f)
```

```{r}
#| label: fig-indices-model-evenness
#| fig-scap: "Evenness (Part b, Figure 3)"
#| fig-cap: !expr glue("Evenness pr. subsection. Shared letters indicate non-significant differences among estimated marginal means according to a compact letter display based on Tukey-adjusted pairwise comparisons at the 5% significance level.")
#| echo: false
#| output: true
#| message: false
#| warning: false


subsection_indices$year <- factor(subsection_indices$year, levels = years)
evenness_emm_df <- as.data.frame(evenness_emmeans_f)
evenness_emm_df$year <- factor(evenness_emm_df$year, levels = years)

evenness_cld <- cld(
  evenness_emmeans_f,
  adjust = "tukey",
  Letters = letters
)

evenness_cld <- as.data.frame(evenness_cld)
evenness_cld$year <- factor(evenness_cld$year, levels = years)

cl_cols <- intersect(c("asymp.LCL", "lower.CL"), names(evenness_emm_df))[1]
cu_cols <- intersect(c("asymp.UCL", "upper.CL"), names(evenness_emm_df))[1]

plot_evenness <- ggplot() +
  geom_jitter(
    data = subsection_indices,
    aes(x = year, y = evenness),
    width = jitter_width,
    alpha = 0.4,
    color = "darkgreen",
    size = 2
  ) +
      geom_crossbar(data = evenness_emm_df,
              aes(x = year, y = emmean,
                  ymin = .data[[cl_cols]], 
                  ymax = .data[[cu_cols]]),
              width = crossbar_width,
              fill = "black", 
              alpha = 0.1,      # <<< no fill
              colour = "black",   # <<< outline back
              linewidth = 0.2)  +  # <<< very thin
  # geom_point(data = evenness_emm_df,
  #            aes(x = year, y = emmean),
  #            size = 7, colour = "black", shape = 95) +
  geom_text(
    data = evenness_cld,
    aes(x = year, y = max(subsection_indices$evenness), label = .group),
    size = 3,
    color = "black",
    inherit.aes = FALSE, nudge_x = 0.2
  ) +
  labs(x = "", y = "Evenness")

plot_evenness
```
```{r}
#| label: indices-tbl-evenness
#| echo: false
#| output: true
#| message: false
#| warning: false


cld_table(evenness_cld, "Estimated marginal means and compact letter display for evenness per survey year")
```

### Shannon

```{r}
#| label: indices-model-shannon
#| echo: false
#| output: false
#| message: false
#| warning: false

model_shannon_sub_f <- lmer(
  shannon ~ factor(year) + (1 | subsection),
  data = subsection_indices
)

summary(model_shannon_sub_f)

shannon_emmeans_f <- emmeans(model_shannon_sub_f, ~ year) 

shannon_pairs_f <- emmeans(model_shannon_sub_f, ~ year) |> 
  pairs()

shannon_emmeans_f
shannon_pairs_f

shannon_pairs_df <- as.data.frame(shannon_pairs_f)
```

```{r}
#| label: fig-indices-model-shannon
#| fig-scap: "Shannon (Part c, Figure 3)"
#| fig-cap: !expr glue("Shannon diversity pr. year calculate pr subsection. Shared letters indicate non-significant differences among estimated marginal means according to a compact letter display based on Tukey-adjusted pairwise comparisons at the 5% significance level.")
#| echo: false
#| output: true
#| message: false
#| warning: false

subsection_indices$year <- factor(subsection_indices$year, levels = years)
shannon_emm_df <- as.data.frame(shannon_emmeans_f)
shannon_emm_df$year <- factor(shannon_emm_df$year, levels = years)

shannon_cld <- cld(
  shannon_emmeans_f,
  adjust = "tukey",
  Letters = letters
)

shannon_cld <- as.data.frame(shannon_cld)
shannon_cld$year <- factor(shannon_cld$year, levels = years)

cl_cols <- intersect(c("asymp.LCL", "lower.CL"), names(shannon_emm_df))[1]
cu_cols <- intersect(c("asymp.UCL", "upper.CL"), names(shannon_emm_df))[1]

plot_shannon <- ggplot() +
  geom_jitter(
    data = subsection_indices,
    aes(x = year, y = shannon),
    width = jitter_width,
    alpha = 0.4,
    color = "darkgreen",
    size = 2
  ) +
        geom_crossbar(data = shannon_emm_df,
              aes(x = year, y = emmean,
                  ymin = .data[[cl_cols]], 
                  ymax = .data[[cu_cols]]),
              width = crossbar_width,
              fill = "black", 
              alpha = 0.1,      # <<< no fill
              colour = "black",   # <<< outline back
              linewidth = 0.2)  +  # <<< very thin

  # geom_point(data = shannon_emm_df,
  #            aes(x = year, y = emmean),
  #            size = 7, colour = "black", shape = 95) +
  geom_text(
    data = shannon_cld,
    aes(x = year, y = max(subsection_indices$shannon), label = .group),
    size = 3,
    color = "black",
    inherit.aes = FALSE, nudge_x = 0.2
  ) +
  labs(x = "", y = "Shannon")

plot_shannon
```
```{r}
#| label: indices-tbl-shannon
#| echo: false
#| output: true
#| message: false
#| warning: false


cld_table(shannon_cld, "Estimated marginal means and compact letter display for evenness per survey year")
```
### Turnover

```{r}
#| label: data-indices-turnover-calculation
#| echo: false
#| output: false
#| message: false
#| warning: false

# Check number of unique years per plot
plot_years <- subsection_species |>
  group_by(subsection) |>
  summarise(n_years = n_distinct(year), .groups = "drop")

# Filter plots with at least 2 years (turnover requires temporal comparison)
valid_plots <- plot_years |>
  filter(n_years > 1) |>
  pull(subsection)

# Filter data to valid plots only
species_long_filtered <- subsection_species |>
  filter(subsection %in% valid_plots)

# Calculate turnover per plot between years
turnover_results <- turnover(
  df = species_long_filtered,
  time.var = "year",
  species.var = "taxon_code",
  abundance.var = "foo",
  replicate.var = "subsection",
  metric = "total"
)

```

```{r}
#| label: indices-model-turnover
#| echo: false
#| output: false
#| message: false
#| warning: false

m_turnover_sub_f <- lmer(total ~ factor(year) + (1|subsection),
                         data = turnover_results)

summary(m_turnover_sub_f)

turnover_emmeans <- emmeans(m_turnover_sub_f, ~ year) |>
  pairs()

turnover_emmeans

```

```{r}
#| label: fig-indices-model-turnover
#| fig-scap: "Turnover (Part d, Figure 3)"
#| fig-cap: !expr glue("Turnover pr. year calculate pr subsection, but only pr present all survey years. A totalt of {length(unique(turnover_results$subsection))} subsections maintain more than 1 plots for all 4 surveys. Shared letters indicate non-significant differences among estimated marginal means according to a compact letter display based on Tukey-adjusted pairwise comparisons at the 5% significance level.")
#| echo: false
#| output: true
#| message: false
#| warning: false

## 2️⃣  Define the canonical year factor (four ticks)
all_years <- factor(c("2007","2012","2017","2022"),
                    levels = c("2007","2012","2017","2022"))


## 3️⃣  Marginal means for the three intervals that really exist
emm_raw_turnover <- emmeans(m_turnover_sub_f, ~ year) |>          # returns 3 rows
  as.data.frame() |>
  mutate(year = factor(year, levels = levels(all_years)))   # align factor


## 4️⃣  Build the *plot* data frame – pad the missing 2007 row

turnover_plot_df <- tibble(year = all_years) |>          # 4 rows, ordered
  left_join(emm_raw_turnover, by = "year") |>                    # bring in the 3 real rows
  mutate(x_pos = as.integer(year))                       # 1‑4 for the axis


## 5️⃣  Compact‑letter‑display (add a dummy row for 2007)
# CLD for the three real means
cld_three <- cld(emmeans(m_turnover_sub_f, ~ year),
                 adjust = "tukey", Letters = letters) |>
  as.data.frame() |>
  mutate(year = factor(year, levels = levels(all_years)),
         x_pos = as.integer(year))

# Dummy row for 2007 (no label)
cld_dummy <- data.frame(year   = factor("2007", levels = levels(all_years)),
                        .group = "",               # empty string → no letter
                        x_pos  = 1,
                        stringsAsFactors = FALSE)

# Combine – now we have four rows, matching turnover_plot_df
turnover_cld <- bind_rows(cld_dummy, cld_three) |>
  arrange(x_pos)

turnover_jitter <- turnover_results |>
  filter(year %in% c(2012, 2017, 2022))

## 6️⃣  Plot – everything comes from the two tidy tables

plot_turnover <- ggplot() +
  geom_jitter(data = turnover_jitter,
              aes(x = as.integer(factor(year, levels = levels(all_years))),
                  y = total),
              width = jitter_width, alpha = 0.4, colour = "darkgreen", size = 2) +
  geom_crossbar(data = turnover_plot_df,
                aes(x = x_pos, y = emmean,
                    ymin = .data[[cl_cols]],
                    ymax = .data[[cu_cols]]),
              width = crossbar_width,
              fill = "black", 
              alpha = 0.1,      # <<< no fill
              colour = "black",   # <<< outline back
              linewidth = 0.2)  +  # <<< very thin
  # geom_point(data = turnover_plot_df,
  #            aes(x = x_pos, y = emmean),
  #            size = 7, colour = "black", shape = 95) +
  geom_text(data = turnover_cld,
            aes(x = x_pos, label = .group),
            y = max(turnover_jitter$total, na.rm = TRUE),
            size = 3, colour = "black", nudge_x = 0.2) +
  scale_x_continuous(name = "",
                     breaks = 1:4,
                     labels = c("2007", "2012", "2017", "2022")) +
  labs(y = "Turnover (total)")

plot_turnover
```
```{r}
#| label: indices-tbl-turnover
#| echo: false
#| output: true
#| message: false
#| warning: false


cld_table(turnover_cld, "Estimated marginal means and compact letter display for evenness per survey year")
```

### Plots combined
````{r}
#| label: fig-indices-wrapped
#| fig-scap: "All indices plot, wrapped (Figure 3)"
#| fig-cap: !expr glue("Diversity metrices, a) Shannon diversity, b) Species richness, c) Evenness, c) Species turnover. Shared letters indicate non-significant differences among estimated marginal means according to a compact letter display based on Tukey-adjusted pairwise comparisons at the 5% significance level. (Figure 3)")
#| echo: false
#| output: true
#| message: false
#| warning: false

grid_labeled <- (
  (plot_richness | plot_evenness) /
  (plot_shannon  | plot_turnover)
) +
  plot_layout(
    guides  = "collect",
    heights = c(1, 1),
    widths  = c(1, 1)
  ) +
  plot_annotation(
    tag_levels = "a",
    tag_suffix = ")"
  ) &
  theme(legend.position = "bottom", plot.tag = element_text(size = fig_text_size))

grid_labeled

ggsave("figures/fig03.pdf", grid_labeled, device = cairo_pdf,
       width = 174, height = 120, units = "mm")
#normalizePath("figures/fig03.pdf")    # prints the exact location

````
```{r}
#| label: indices-tbl
#| echo: false
#| output: true
#| message: false
#| warning: false



```

{{< pagebreak >}}

## Community composition

```{r}
#| label: importing-data-nmds
#| echo: false
#| output: false
#| message: false
#| warning: false

community_matrix <- subsection_species_wide |> 
  unite("sub_year_vt", subsection, year, veg_type, sep = "_", remove = FALSE) |> 
  as.data.frame() |> 
  column_to_rownames(var = "sub_year_vt") |>
  dplyr::select(-year, -subsection, -veg_type, -no_plots)

meta <- subsection_species_wide |>
  mutate(sub_year_vt = paste(subsection, year, veg_type, sep = "_")) |>
  dplyr::select(sub_year_vt, year, veg_type, subsection) |> 
  distinct()

```
### NMDS
Community composition data ready for analysis consists of `r length(unique(meta$subsection))` subsections from `r length(unique(meta$year))` years of surveys. `r  ncol(community_matrix)` species have been observed across all surveys.

```{r}
#| label: nmds-run
#| echo: false
#| output: false
#| message: false
#| warning: false


# Set seed for reproducibility
set.seed(42)

# Run NMDS on presence/absence community matrix
# nmds_result <- metaMDS(community_matrix, k = 3, trymax = 100)
# write_rds(nmds_result, "nmds_result")

nmds_result <- read_rds("nmds_result")

print(nmds_result)

nmds_scores <- as.data.frame(scores(nmds_result, display = "sites"))

#nrow(community_matrix) # to check that they are the same lenght
#nrow(nmds_scores) 

# Extract NMDS scores for plots
nmds_scores <- as.data.frame(scores(nmds_result, display = "sites"))
nmds_scores$sub_year_vt <- rownames(nmds_scores)

write_rds(nmds_result,"nmds_results")
write_rds(nmds_scores,"nmds_scores")
```

```{r}
#| label: fig-nmds
#| fig-scap: "NMDS (Figure 4)"
#| fig-cap: !expr glue("Plant community dynamics. Non-metric multidimensional scaling (NMDS) ordination of vegetation composition across years and vegetation types. Each panel represents subset of years of the full NMDS including all data to keep NMDS distances comparable. Grey lines connect plots within vegetation types, the within-type centroid distances. Black dashed lines indicate distances between vegetation-type centroids.")
#| echo: false
#| output: true
#| message: false
#| warning: false

nmds_plot_data <- nmds_scores |>
  left_join(meta, by = "sub_year_vt") |>
  mutate(year = factor(year))  # convert year to factor for plotting

# Function to compute convex hull safely
find_hull <- function(df) {
  df <- na.omit(df)
  if (nrow(df) < 3) return(df)
  df[chull(df$NMDS1, df$NMDS2), ]
}

# Compute hulls per year and veg_type
hulls <- nmds_plot_data |>
  group_by(year, veg_type) |>
  do(find_hull(.))

# Close hull polygons
hulls_closed <- hulls |>
  group_by(year, veg_type) |>
  do({
    df <- .
    if (nrow(df) > 1) df <- rbind(df, df[1, ])
    df
  })

# Compute centroids per year and veg_type
centroids <- nmds_plot_data |>
  group_by(year, veg_type) |>
  summarize(centroid_NMDS1 = mean(NMDS1),
            centroid_NMDS2 = mean(NMDS2),
            .groups = "drop")

centroid_paths <- centroids |>
  arrange(veg_type, year)

centroid_pairs <- centroids |>
  arrange(year, veg_type) |>
  group_by(year) |>
  reframe({
    pairs <- combn(n(), 2)
    tibble(
      x    = centroid_NMDS1[pairs[1,]],
      y    = centroid_NMDS2[pairs[1,]],
      xend = centroid_NMDS1[pairs[2,]],
      yend = centroid_NMDS2[pairs[2,]]
    )
  })

# Create separate plots for each year
lims_x <- base::range(nmds_plot_data$NMDS1, na.rm = TRUE)
lims_y <- base::range(nmds_plot_data$NMDS2, na.rm = TRUE)

centroids$year        <- as.character(centroids$year)
nmds_plot_data$year   <- as.character(nmds_plot_data$year)
hulls_closed$year     <- as.character(hulls_closed$year)
centroid_pairs$year   <- as.character(centroid_pairs$year)

years <- unique(centroids$year)  # will now be characters


plots <- lapply(years, function(y) {
  
  df       <- nmds_plot_data |> filter(year == y)
  hull_df  <- hulls_closed   |> filter(year == y)
  cent_df  <- centroids      |> filter(year == y)
  
  df_joined <- df |>
    left_join(cent_df, by = c("year", "veg_type"))
  
  # centroid–centroid pairs
  pair_df <- cent_df |>
    arrange(veg_type) |>
    reframe({
      if (n() < 2) return(NULL)
      pairs <- combn(n(), 2)
      tibble(
        x    = centroid_NMDS1[pairs[1,]],
        y    = centroid_NMDS2[pairs[1,]],
        xend = centroid_NMDS1[pairs[2,]],
        yend = centroid_NMDS2[pairs[2,]]
      )
    })
  
  ggplot() +   # ⬅️ IMPORTANT: remove global aes()
    
    # Hulls
    geom_polygon(data = hull_df,
                 aes(NMDS1, NMDS2, fill = veg_type, group = veg_type),
                 alpha = 0.2,
                 colour = NA) +
    
    geom_path(data = hull_df,
              aes(NMDS1, NMDS2, group = veg_type),
              colour = "white",
              linewidth = 0.3) +
    
    # 🔹 centroid–centroid
    geom_segment(data = pair_df,
                 aes(x = x, y = y, xend = xend, yend = yend),
                 linetype = "dashed",
                 linewidth = 0.7,
                 colour = "black") +
    
    # 🔹 plot–centroid
    geom_segment(data = df_joined,
                 aes(x = NMDS1, y = NMDS2,
                     xend = centroid_NMDS1,
                     yend = centroid_NMDS2),
                 linewidth = 0.3,
                 colour = "grey40",
                 alpha = 0.6) +
    
    # Observations
    geom_point(data = df,
               aes(NMDS1, NMDS2,
                   shape = veg_type,
                   colour = veg_type),
               size = 2,
               alpha = 0.7) +
    
    # Centroids
    geom_point(data = cent_df,
               aes(centroid_NMDS1,
                   centroid_NMDS2,
                   fill = veg_type),
               shape = 21,
               size = 3,
               colour = "black") +
    
geom_label(data = cent_df,
           aes(centroid_NMDS1, centroid_NMDS2, label = veg_type),
           vjust = -0.2, hjust = 1.3, size = 2.5,
           fill = "white", alpha = 0.7,
           label.size = 0,                      # no border line
           label.padding = unit(0.1, "lines"))+  # tight padding
    coord_cartesian(xlim = lims_x,
                    ylim = lims_y) +
    annotate("text",
         x = lims_x[1], 
         y = lims_y[2],
         label = y,
         hjust = 0, 
         vjust = 1,
         size = 3)+
    theme(legend.position = "none")
})

# Combine plots in a grid
wrap_plots(plots, ncol = 2)  

ggsave("figures/fig04.pdf", wrap_plots(plots, ncol = 2), device = cairo_pdf,
       width = 174, height = 120, units = "mm")
#normalizePath("figures/fig04.pdf")    # prints the exact location

```

### Between vegetation composition
```{r}
#| label: nmds-between-analysis
#| echo: false
#| output: false
#| warning: false
#| message: false

# Pairwise distances between vegetation type centroids
  centroid_pairs <- centroids |>
    inner_join(centroids, by = "year", suffix = c("1", "2")) |>
    filter(as.character(veg_type1) < as.character(veg_type2)) |>
    mutate(
      dist = sqrt(
        (centroid_NMDS11 - centroid_NMDS12)^2 +
          (centroid_NMDS21 - centroid_NMDS22)^2
      ),
      veg_pair = interaction(veg_type1, veg_type2, drop = TRUE)
    ) |>
    dplyr::select(year, veg_type1, veg_type2, veg_pair, dist)
  
  # Summary table
  centroid_dist_summary <- centroid_pairs |>
    group_by(year) |>
    summarise(
      mean_dist   = mean(dist),
      median_dist = median(dist),
      sd_dist     = sd(dist),
      n_pairs     = n(),
      .groups = "drop"
    )

# Mixed model
m_convergence <- lmer(
  dist ~ factor(year) + (1 | veg_pair),
  data = centroid_pairs
)
summary(m_convergence)
anova(m_convergence)

# EMMs and pairwise contrasts
emm_year <- emmeans(m_convergence, ~ year)
pairs(emm_year)

emm_df <- as.data.frame(emm_year) |>
  mutate(
    year  = factor(year, levels = levels(emm_year@grid$year)),
    x_pos = as.integer(year)
  )

# CLD
centroid_cld <- cld(emm_year, adjust = "tukey", Letters = letters) |> 
  as.data.frame() |>
  mutate(
    year  = factor(year, levels = levels(emm_year@grid$year)),
    x_pos = as.integer(year)
  )

```

```{r}
#| label: nmds-between-analysis-print
#| echo: false
#| output: true
#| warning: false
#| message: false

# Model structure (one line description)
cat("Model: dist ~ factor(year) + (1 | veg_pair), fitted by REML\n")
cat("N observations:", nobs(m_convergence), 
    "| N groups (veg_pair):", ngrps(m_convergence), "\n\n")

# Model fixed effects
broom.mixed::tidy(m_convergence, effects = "fixed") |>
  dplyr::select(term, estimate, std.error, statistic) |>
  dplyr::mutate(term = stringr::str_replace(term, "factor\\(year\\)", "Year ")) |>
  knitr::kable(digits = 3, caption = "Fixed effects: centroid distance ~ survey year (factor)")


# 1. EMMs per survey with 95% CI
emm_df |>
  dplyr::select(year, emmean, SE, lower.CL, upper.CL) |>
  knitr::kable(digits = 3, caption = "Estimated mean centroid distances per survey")

# 2. Pairwise comparisons with p-values
pairs(emm_year, adjust = "sidak") |>
  as.data.frame() |>
  dplyr::select(contrast, estimate, SE, t.ratio, p.value) |>
  knitr::kable(digits = 3, caption = "Pairwise contrasts between surveys (Sidak-adjusted)")

```

```{r}
#| label: fig-nmds-between
#| fig-scap: "Between community dynamics (Figure 5)"
#| fig-cap: !expr glue("Between plant community dynamics. Estimated mean distance between vegetation-type centroids within each survey year based on linear mixed-effects models. Black lines are estimated means (± SE), and grey dots represent raw subsection-level distances (NMDS units). Shared letters indicate non-significant differences among estimated marginal means according to a compact letter display based on Tukey-adjusted pairwise comparisons at the 5% significance level. (Figure 5)")
#| echo: false
#| output: true
#| message: false
#| warning: false


y_max <- max(
  max(emm_df$upper.CL, na.rm = TRUE),
  max(centroid_pairs$dist, na.rm = TRUE)
) * 1.05

between <- ggplot() +
  geom_jitter(data = centroid_pairs,
              aes(x = as.integer(factor(year)), y = dist),
              width = jitter_width, height = 0,
              alpha = 0.4, colour = "darkgreen", , size = 2) +
  geom_crossbar(data = emm_df,
              aes(x = x_pos, y = emmean, ymin = lower.CL, ymax = upper.CL),
              width = crossbar_width,
              fill = "black", alpha = 0.1,      # <<< no fill
              colour = "black",   # <<< outline back
              linewidth = 0.2)  +  # <<< very thin
  # geom_errorbar(data = emm_df,
  #               aes(x = x_pos, ymin = lower.CL, ymax = upper.CL),
  #               width = 0.2, colour = "black") +
  # geom_point(data = emm_df,
  #            aes(x = x_pos, y = emmean),
  #            size = 7, colour = "black", shape = 95) +
  geom_text(data = centroid_cld,
            aes(x = x_pos, label = .group),
            y = y_max,
            vjust = 0, size = 4, colour = "black",
            nudge_x = 0.2) +
  scale_x_continuous(name = "",
                     breaks = 1:4,
                     labels = c("2007", "2012", "2017", "2022")) +
  coord_cartesian(ylim = c(NA, y_max * 1.05)) +
  labs(y = "Centroid distance (NMDS units)")

between

ggsave("figures/fig05.pdf", between, device = cairo_pdf,
       width = 174, height = 120, units = "mm")
#normalizePath("figures/fig05.pdf")    # prints the exact location
```
```{r}
#| label: tbl-between
#| echo: false
#| output: true
#| message: false
#| warning: false


cld_table(centroid_cld, "Estimated mean centroid distances and compact letter display per survey year")
```

### Within vegetation composition

```{r}
#| label: nmds-analysis-within-data-prep
#| echo: false
#| output: false
#| message: false
#| warning: false

## 1️⃣  Global NMDS site scores + metadat

nmds_result <- read_rds("nmds_results")
# # 1.1  Pull the three NMDS axes for every plot
# within_site_scores <- scores(nmds_result, display = "sites") |>
#   as.data.frame() |>
#   tibble::rownames_to_column("sub_year_vt")
# 
# # 1.2  Join the metadata (veg_type, year, etc.)
# within_site_scores <- within_site_scores |>
#   left_join(meta, by = "sub_year_vt")

# 1.3  Keep only the columns we need
within_site_scores <- scores(nmds_result, display = "sites") |>
  as.data.frame() |>
  tibble::rownames_to_column("sub_year_vt") |>
  left_join(meta, by = "sub_year_vt") |>
  dplyr::select(
    sub_year_vt,
    veg_type,
    year,
    subsection,          # <<< keep this column
    NMDS1, NMDS2, NMDS3 # keep the three NMDS axes
  )

# 1. Compute global centroids (veg_type × year)
within_centroids <- within_site_scores |>
  dplyr::group_by(veg_type, year) |>
  dplyr::summarise(
    cen1 = mean(NMDS1),
    cen2 = mean(NMDS2),
    cen3 = mean(NMDS3),
    .groups = "drop"
  )

# 2. Join and create the distance column (single operation)
within_site_scores <- within_site_scores |>
  dplyr::left_join(within_centroids,
                   by = c("veg_type", "year")) |>
  dplyr::mutate(
    dist_to_centroid = sqrt(
      (NMDS1 - cen1)^2 +
        (NMDS2 - cen2)^2 +
        (NMDS3 - cen3)^2
    )
  ) |>
  # <‑‑ make sure it is numeric
  dplyr::mutate(dist_to_centroid = as.numeric(dist_to_centroid)) |>
  dplyr::select(-cen1, -cen2, -cen3)
##  Global y‑axis limits (computed once)
# Raw distances from the full data set (used for the jitter points)
global_y_max <- within_site_scores |>
  dplyr::summarise(max_dist = max(dist_to_centroid, na.rm = TRUE)) |>
  dplyr::pull(max_dist)

# Upper bound for the error‑bar + mean point (adds a small cushion)
global_y_upper <- global_y_max * 1.10   # 10 % headroom for the CLD letters
global_y_limits <- c(0, global_y_upper)   # lower bound = 0 (distances cannot be negative)

```


```{r}
#| label: nmds-analysis-within
#| echo: false
#| output: false
#| message: false
#| warning: false


veg_vec   <- unique(within_site_scores$veg_type)
within_plots <- vector("list", length(veg_vec))
within_fixed <- tibble()   # collect fixed‑effect summaries
within_pairwise  <- tibble()   # pairwise year contrasts for all veg types
within_emm_table <- tibble()

for (i in seq_along(veg_vec)) {
  vt <- veg_vec[i]
  

  ## 2.1  Subset data (year stays a factor for the model)
  vt_dat <- within_site_scores |>
    dplyr::filter(veg_type == vt) |>
    dplyr::mutate(year = factor(year))
  

  ## 2.2  Fit mixed‑effects model (year fixed, subsection random)

  mod <- lmerTest::lmer(
    dist_to_centroid ~ year + (1 | subsection),
    data = vt_dat,
    REML = TRUE
  )
  

  ## 2.3  EMMeans (needed for both CLD and pairwise contrasts)

  emm <- emmeans::emmeans(mod, ~ year)
  
  ## 2.31
  within_cld <- multcomp::cld(emm, adjust = "tukey", Letters = letters, sort = FALSE) |>
  tibble::as_tibble() |>
  dplyr::mutate(veg_type = vt, group = trimws(.group)) |>
  dplyr::select(veg_type, year, emmean, SE,
                dplyr::any_of(c("lower.CL", "asymp.LCL")),
                dplyr::any_of(c("upper.CL", "asymp.UCL")), group)

  within_emm_table <- dplyr::bind_rows(within_emm_table, within_cld)

  ## 2.4  CLD letters (Sidak‑adjusted) – for the plot

  cld_df <- multcomp::cld(
    emm,
    adjust = "tukey",
    Letters = letters,
    sort = FALSE
  ) |> as_tibble() |> dplyr::select(year, .group)
  

  ## 2.5  Pairwise year contrasts (the numbers behind the CLD)

  pairwise_cmp <- emmeans::contrast(
    emm,
    method = "pairwise",
    adjust = "sidak"
  ) |> as_tibble() |> dplyr::rename(
    contrast = contrast,
    estimate = estimate,
    SE       = SE,
    df       = df,
    t_ratio  = t.ratio,
    p_adj    = p.value
  ) |> dplyr::mutate(veg_type = vt)
  

  ## 2.6  Append pairwise results to the master table

  within_pairwise <- dplyr::bind_rows(within_pairwise, pairwise_cmp)
  
  ## 2.7  Plot‑ready summary (mean ± SE) + attach CLD letters
  plot_stats <- vt_dat |>
    dplyr::group_by(year) |>
    dplyr::summarise(
      mean_dist = mean(dist_to_centroid),
      se_dist   = sd(dist_to_centroid) / sqrt(n()),
      .groups = "drop"
    ) |>
    dplyr::left_join(cld_df, by = "year")
  
  within_centroid_cld <- plot_stats |>
    dplyr::mutate(
      y_pos = global_y_upper * 0.92   # fixed vertical position inside panel
    ) |>
    dplyr::select(year, .group, y_pos)
  
## 2.8  Build the facet plot
p <- ggplot() +
  geom_jitter(data = vt_dat,
              aes(x = as.integer(factor(year)), y = dist_to_centroid),
              width = jitter_width, height = 0,
              alpha = 0.4, colour = "darkgreen", size = 2) +
  
  geom_crossbar(data = plot_stats,
                aes(x = as.integer(factor(year)),
                    y = mean_dist,
                    ymin = mean_dist - se_dist,
                    ymax = mean_dist + se_dist),
                width = crossbar_width,
                fill = "black", alpha = 0.1,
                colour = "black",
                linewidth = 0.2) +
  
  geom_text(data = within_centroid_cld,
            aes(x = as.integer(factor(year)), label = .group),
            y = global_y_upper * 0.92,
            vjust = 0, size = 4, colour = "black",
            nudge_x = 0.2) +
  
  scale_x_continuous(name = "",
                     breaks = 1:4,
                     labels = c("2007", "2012", "2017", "2022")) +
  coord_cartesian(ylim = global_y_limits) +
  labs(title = vt, y = "Centroid distance (NMDS units)") 
  
  ## 2.9  Store the plot
  within_plots[[i]] <- p
  
  ## 2.10  Extract fixed‑effect summary (includes p‑values)
  fixed_tab <- summary(mod)$coefficients |>
    as_tibble(rownames = "term") |>
    dplyr::rename(
      estimate   = Estimate,
      std.error  = `Std. Error`,
      statistic  = `t value`,
      p.value    = `Pr(>|t|)`
    ) |>
    dplyr::mutate(
      veg_type = vt,
      effect   = "fixed"
    )
  
  ## 2.11  Append to the master MLM table
  within_fixed <- dplyr::bind_rows(within_fixed, fixed_tab)
}

```

```{r}
#| label: fig-nmds-within
#| fig-scap: "Within community dynamics (Figure 6)"
#| fig-cap: !expr glue("Within community dynamics. Estimated mean centroids-subsection distances pr community each survey year based on linear mixed-effects models. Black lines are estimated means (± SE), and grey dots represent raw distances (NMDS units). Shared letters indicate non-significant differences among estimated marginal means according to a compact letter display based on Tukey-adjusted pairwise comparisons at the 5% significance level. (Figure 6)")
#| echo: false
#| output: true
#| message: false
#| warning: false

within_grid_plot <- wrap_plots(within_plots, ncol = 2) &
  theme(plot.margin = margin(1, 1, 1, 1))

# Show the figure
print(within_grid_plot)

ggsave("figures/fig06.pdf", within_grid_plot, device = cairo_pdf,
       width = 174, height = 120, units = "mm")

#normalizePath("figures/fig06.pdf")    # prints the exact location

```

```{r}
#| label: tbl-within
#| echo: false
#| output: true
#| message: false
#| warning: false


within_emm_table |>
  dplyr::arrange(veg_type, year) |>
  knitr::kable(digits = 3,
    caption = "Estimated marginal means and compact letter display per vegetation type and survey year")
```

{{< pagebreak >}}

## Plant functional types

### Functional groups

```{r}
#| label: data-import-pft
#| echo: false
#| output: false
#| message: false
#| warning: false

merged_data <- readRDS("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/nero_analysis/data/merged_data.rds") |> 
  group_by(year, subsection) |> 
  mutate(no_plots = n_distinct(plot_id)) |> 
  ungroup()

pivot <- merged_data |> 
  dplyr::select(year, vt_section,subsection, plot, plot_id, no_plots, veg_type, func_type, ecoveg_gfc, ecoveg_sgfc,presence, taxon_code) |> 
  distinct(year, vt_section,subsection, plot, plot_id, no_plots, veg_type, presence, taxon_code)

meta <- merged_data |> 
  dplyr::select(taxon_code, func_type, ecoveg_gfc, ecoveg_sgfc) |> 
  distinct()

pivot_wide <- pivot |> 
  pivot_wider(names_from = taxon_code, values_from = presence, values_fill = 0 )

species_long <- pivot_wide |> 
  pivot_longer(cols = 8:ncol(pivot_wide), names_to = "taxon_code", values_to = "occurrence")

species_sub <- species_long |> 
  group_by(year,subsection,no_plots,veg_type, taxon_code) |> 
  summarise(count = sum(occurrence)) |> 
  mutate(fraction = count/no_plots) |> 
  left_join(meta, by = "taxon_code")

```


```{r}
#| label: model-pft
#| echo: false
#| output: false
#| message: false
#| warning: false

func_summed <- species_sub |> 
  group_by(year, subsection, veg_type, fraction, func_type) |> 
  reframe(sum_frac = sum(fraction)) |> 
  filter(func_type != "rock")

func_types     <- unique(func_summed$func_type)
plot_list_pft  <- list()
fixed_list_pft <- list()
emm_list_pft   <- list()
emm_cld_list_pft <- list()
pairs_list_pft <- list()
has_diff_groups_pft <- logical(0)   # separate from the species one

for(pft in func_types){
  
  df_pft <- func_summed |>
    dplyr::filter(func_type == pft)
  
  if(nrow(df_pft) < 5) next
  
  # ----- MODEL -----
  mod <- lmer(
    sum_frac ~ factor(year) + (1 | subsection),
    data = df_pft,
    REML = TRUE
  )
  fixed_list_pft[[pft]] <- mod
  
  # ----- PAIRWISE TESTS -----
  pairwise <- mod |>
    emmeans(~ factor(year)) |>
    pairs(adjust = "tukey")
  pairs_list_pft[[pft]] <- pairwise
  
  # ----- EMMEANS -----
  emm_df <- mod |>
    emmeans(~ factor(year)) |>
    as.data.frame() |>
    dplyr::rename_with(~ "year", 1) |>
    dplyr::mutate(year = as.numeric(as.character(year)))
  emm_list_pft[[pft]] <- emm_df

  
  # ----- CLD LETTERS -----
  cld_years <- mod |>
    emmeans(~ factor(year)) |>
    multcomp::cld(adjust = "tukey", Letters = letters) |>
    as.data.frame() |>
    dplyr::rename_with(~ "year", 1) |>
    dplyr::mutate(year = as.numeric(as.character(year)))

  cld_labels <- cld_years |>
    dplyr::select(year, .group)
  
  year_emm_lab <- emm_df |>
    dplyr::left_join(cld_labels, by = "year")
  
  has_diff_groups_pft[[pft]] <- n_distinct(trimws(year_emm_lab$.group)) > 1
  
  # ----- PLOT -----
  p <- ggplot() +
    geom_jitter(
      data = df_pft,
      aes(x = year, y = sum_frac),
      width = jitter_width, alpha = 0.5, color = "darkgreen", , size = 2) +
    geom_crossbar(
      data = emm_df,
      aes(x = year, y = emmean, ymin = lower.CL, ymax = upper.CL),
      width = crossbar_width, fill = "black", alpha = 0.1,
      colour = "black", linewidth = 0.2
    ) +
    # geom_point(
    #   data = emm_df,
    #   aes(x = year, y = emmean),
    #   size = 7, colour = "black", shape = 95
    # ) +
    geom_text(
      data = year_emm_lab,
      aes(x = year, label = .group), y = max(df_pft$sum_frac, na.rm = TRUE),
      size = 3, color = "black", inherit.aes = FALSE, nudge_x = 0.45
    ) +
    labs(x = "", y = "Summed FOO") +
    annotate("text", x = -Inf, y = Inf, label = pft,
             hjust = -0.1, vjust = 1.4, size = 3)+
    scale_x_continuous(name = "",
                   breaks = c(2007, 2012, 2017, 2022),
                   labels = c("2007", "2012", "2017", "2022"))+
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.18)))
  
  plot_list_pft[[pft]] <- p
}

emm_cld_list_pft[[pft]] <- year_emm_lab

ggsave("figures/fig07.pdf", plot_list_pft[[pft]], device = cairo_pdf,
       width = 174, height = 120, units = "mm")
#normalizePath("figures/fig07.pdf")    # prints the exact location

```


```{r}
#| label: fig-pft
#| fig-scap: "Plant functional groups (Figure 7)"
#| fig-cap: !expr glue("Plant functional group dynamics. Summed frequency of occurrence (FOO) for plant functional groups. Shared letters indicate non-significant differences among estimated marginal means according to a compact letter display based on Tukey-adjusted pairwise comparisons at the 5% significance level. (Figure 7)")
#| echo: false
#| output: true
#| warning: false
#| message: false

grid_pft <- wrap_plots(plot_list_pft, ncol = 2) +
  #plot_annotation(
  #  tag_levels = "a",
  #  tag_suffix = ")"
  #) &
  theme(plot.tag = element_text(size = fig_text_size))

grid_pft


```

```{r}
#| label: table-pft
#| echo: false
#| output: true
#| warning: false
#| message: false

# Fixed effects table
# purrr::map_dfr(fixed_list_pft, broom.mixed::tidy, effects = "fixed", .id = "func_type") |>
#   dplyr::relocate(func_type) |>
#   knitr::kable(digits = 3, caption = "Fixed effects per functional type")

suppressMessages(purrr::imap_dfr(fixed_list_pft, function(mod, pft) {
  mod |>
    emmeans(~ factor(year)) |>
    multcomp::cld(adjust = "tukey", Letters = letters) |>
    as.data.frame() |>
    dplyr::rename_with(~ "year", 1) |>
    dplyr::mutate(
      year      = as.numeric(as.character(year)),
      func_type = pft,
      group     = trimws(.group)
    ) |>
    dplyr::select(func_type, year, emmean, SE, df, lower.CL, upper.CL, group)
}) |>
  dplyr::arrange(func_type, year) |>
  knitr::kable(
    digits = 3,
    caption = "Estimated marginal means and compact letter display (based adjusted pairwise comparisons at the 5% significance level) per functional type and survey year"
  ))

```
{{< pagebreak >}}

### Individual species

```{r}
#| label: species-data-prep
#| echo: false
#| output: false
#| warning: false
#| message: false

species_taxon_code <- readRDS("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/nero_analysis/data/taxon_code_species_key.rds")

species_sub <- readRDS("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/nero_analysis/data/species_sub.rds") |> 
  filter(taxon_code != "rock") |> 
  left_join(species_taxon_code, by = "taxon_code")


```

```{r}
#| label: species-analysis
#| echo: false
#| output: true
#| warning: false
#| message: false

species_list <- unique(species_sub$taxon_code)
model_list      <- list()
pairwise_list   <- list()
emm_list        <- list()
plot_list       <- list()
has_diff_groups <- logical(0)

for(sp in species_list){
  
  df_sp <- species_sub |>
    dplyr::filter(taxon_code == sp)
  
  if(nrow(df_sp) < 5) next
  
  # ----- MODEL -----
  mod <- suppressMessages(lmer(
    fraction ~ factor(year) + (1 | subsection),
    data = df_sp,
    REML = TRUE
  ))
  model_list[[sp]] <- mod
  
  # ----- EMMEANS (fit once, reuse) -----
  emm_obj <- suppressMessages(emmeans(mod, ~ factor(year)))
  
  # ----- PAIRWISE TESTS -----
  pairwise_list[[sp]] <- emm_obj |> pairs(adjust = "tukey")
  
  # ----- EMMEANS -----
  emm_df <- emm_obj |>
    as.data.frame() |>
    dplyr::rename_with(~ "year", 1) |>
    dplyr::mutate(year = as.numeric(as.character(year)))
  emm_list[[sp]] <- emm_df
  
  # ----- CLD LETTERS -----
  cld_labels <- suppressMessages(emm_obj |>
    multcomp::cld(adjust = "tukey", Letters = letters) |>
    as.data.frame() |>
    dplyr::rename_with(~ "year", 1) |>
    dplyr::mutate(year = as.numeric(as.character(year))) |>
    dplyr::select(year, .group))
  
  year_emm_lab <- emm_df |>
    dplyr::left_join(cld_labels, by = "year")
  
  has_diff_groups[[sp]] <- n_distinct(trimws(year_emm_lab$.group)) > 1
  
  # ----- PLOT -----
  p <- ggplot() +
    geom_jitter(
      data = df_sp,
      aes(x = year, y = fraction),
      width = jitter_width, alpha = 0.5, color = "darkgreen", size = 2
    ) +
    geom_crossbar(
      data = emm_df,
      aes(x = year, y = emmean, ymin = lower.CL, ymax = upper.CL),
      width = crossbar_width, fill = "black", alpha = 0.1,
      colour = "black", linewidth = 0.2
    ) +
    # geom_point(
    #   data = emm_df,
    #   aes(x = year, y = emmean),
    #   size = 7, colour = "black", shape = 95
    # ) +
    geom_text(
      data = year_emm_lab,
      aes(x = year, label = .group),
      y = max(df_sp$fraction, na.rm = TRUE),
      size = 3, color = "black", inherit.aes = FALSE,
      nudge_x = 0.2
    )+
    labs(x = "", y = "Frequency of occurrence") +
    #annotate("text", x = -Inf, y = Inf, label = sp,
    #         hjust = -0.1, vjust = 1.1, size = 3) +
    scale_x_continuous(name = "",
                       breaks = c(2007, 2012, 2017, 2022),
                       labels = c("2007", "2012", "2017", "2022"))
  
  plot_list[[sp]] <- p
}


```

A total of `r length(model_list)` species were analysed. 
`r sum(has_diff_groups)` of them had any significant differences in means between survey years. Plots of the frequency of occurrence for these `r sum(has_diff_groups)` species are rendered below. 

```{r}
#| label: species-plots
#| echo: false
#| output: asis
#| warning: false
#| message: false
#| results: asis


for(sp in names(plot_list[has_diff_groups])){
  
  species_name <- as.character(unique(species_sub$species[species_sub$taxon_code == sp]))[1]
  
  fig_id   <- gsub("[^a-zA-Z0-9]", "-", sp)
  
  if(sp == "betnan"){
    fig_scap <- paste0("Frequency of occurrence for ", species_name, " (Figure 8)")
  } else {
    fig_scap <- paste0("Frequency of occurrence for \\textit{", species_name, "}")
  }
  
  cat("::: {#fig-", fig_id, " fig-scap='", fig_scap, "'}\n\n", sep = "")
  
  print(plot_list[[sp]])
  
  cat("\n\nFrequency of occurrence for *", species_name, "* across survey years. Points show raw observations, crossbars show estimated marginal means ± 95% CI. Compact letter display indicates statistically significant differences between years (adjusted).\n", sep = "")
  
  cat(":::\n\n")
}

ggsave("figures/fig08.pdf", plot_list[["betnan"]], device = cairo_pdf,
       width = 174, height = 120, units = "mm")
#normalizePath("figures/fig08.pdf")

```

```{r}
#| label: species-tbl-betnan
#| echo: false
#| output: true
#| warning: false
#| message: false

model_list[["betnan"]] |>
  emmeans(~ factor(year)) |>
  multcomp::cld(adjust = "tukey", Letters = letters) |>
  as.data.frame() |>
  dplyr::rename_with(~ "year", 1) |>
  dplyr::mutate(year  = as.numeric(as.character(year)),
                group = trimws(.group)) |>
  dplyr::select(year, emmean, SE, lower.CL, upper.CL, group) |>
  dplyr::arrange(year) |>
  knitr::kable(digits = 3,
    caption = "Estimated marginal means and compact letter display for *Betula nana* per survey year")

```

```{r}
#| label: fig-species-tables
#| echo: false
#| output: false
#| warning: false
#| message: false


```

{{< pagebreak >}}

# Phenology (discussion)

## NDVI
Data available from https://data.g-e-m.dk [@gem2026].

```{r}
#| label: ndvi-data-import
#| echo: false
#| output: false
#| warning: false
#| message: false

ndvi <- readRDS("/Users/ibdj/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/r_generel/ndvi-comparison/ndvi-comparison/biobasis_ndvi.rds") |>   
  clean_names() |> 
  mutate(month = month(date), year = year(date), plot_id = factor(paste0(species, plot)),
         year_c = year - mean(year)) |> 
  filter(ndvi != -9999)
```

```{r}
#| label: ndvi-data-model
#| echo: false
#| output: false
#| warning: false
#| message: false

# model_ndvi <- lmer(
#   ndvi ~ year_c + (1 | plot_id/section),
#   data = ndvi
# )

#write_rds(model_ndvi, "model_ndvi.rds")
model_ndvi <- read_rds("model_ndvi.rds")

# Run once, then comment out:
#boot_ci <- confint(model_ndvi, method = "boot", nsim = 1000)

#write_rds(boot_ci, "boot_ci_ndvi.rds")

# Read cached result every render:
boot_ci <- read_rds("boot_ci_ndvi.rds")

# Fixed effects with CIs
ci <- boot_ci |>
  as.data.frame() |>
  rownames_to_column("term") |>
  rename(lower = `2.5 %`, upper = `97.5 %`)

tidy(model_ndvi, effects = "fixed") |>
  left_join(ci, by = "term") |>
  dplyr::select(term, estimate, std.error, statistic, lower, upper) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  knitr::kable(caption = "Fixed effects")

# Random effects variance components
tidy(model_ndvi, effects = "ran_pars") |>
  dplyr::select(group, term, estimate) |>
  mutate(estimate = round(estimate, 4)) |>
  knitr::kable(caption = "Random effects (variance components)")

ndvi_caption <- glue(
  "NDVI change over time. Mean NDVI per plot per year, based on in situ measurements ",
  "from {min(ndvi$year)} to {max(ndvi$year)}. The trend line shows a significant ",
  "increase of {round(fixef(model_ndvi)['year_c'], 4)} NDVI units per year ",
  "(95% CI: {round(boot_ci['year_c', '2.5 %'], 4)}–",
  "{round(boot_ci['year_c', '97.5 %'], 4)}), representing a cumulative increase of ",
  "{round(fixef(model_ndvi)['year_c'] * (max(ndvi$year) - min(ndvi$year)), 4)} ",
  "NDVI units over the study period."
)

```

```{r}
#| label: fig-ndvi-visualisering
#| fig-scap: "NDVI 2018 - 2024 (Background - discussion)"
#| fig-cap: !expr ndvi_caption
#| echo: false
#| output: true
#| warning: false
#| message: false

# Get predicted values + CIs
mean_year <- mean(ndvi$year)
preds <- ggpredict(model_ndvi, terms = "year_c [all]") |>
  mutate(year = x + mean_year)

# Plot-level yearly means for jitter (cleaner than raw weekly points)
ndvi_means <- ndvi |>
  group_by(plot_id, year) |>
  summarise(ndvi_mean = mean(ndvi), .groups = "drop")

ggplot() +
  geom_jitter(data = ndvi_means,
              aes(x = year, y = ndvi_mean),
              width = jitter_width, height = 0,
              alpha = 0.4, colour = "darkgreen", size = 2) +
  geom_ribbon(data = preds,
              aes(x = year, ymin = conf.low, ymax = conf.high),
              fill = "black", alpha = 0.1) +
  geom_line(data = preds,
            aes(x = year, y = predicted),
            colour = "black", linewidth = 0.2) +
  # geom_point(data = preds |> filter(year %in% unique(ndvi_means$year)),
  #            aes(x = year, y = predicted),
  #            size = 2, colour = "black", shape = 95) +
  scale_x_continuous(name = "",
                     breaks = unique(ndvi_means$year)) +
  labs(y = "NDVI")

```

{{< pagebreak >}}

# References