---
title: "Appendix"
author: "Jacobsen et al 2026"
format:
  pdf:
    fig-width: 6.5  # Approx 16.5cm, slightly wider than text to be safe
    out-width: 100%
    fig-align: center
    toc: true
    fig-pos: "H"
editor: visual
---

# Current climate conditions

Data is available 2007 – 2022. 2008 is the first whole year of temperature data. Data available from GEM. (2020). ClimateBasis Nuuk—Temperature—Air temperature \@ 200 cm—30min average (°C) (Version 1.0) \[Dataset\]. Greenland Ecosystem Monitoring. <https://doi.org/10.17897/PGN3-7597>

## Tempterature

```{r}
#| label: importing-packages
#| output: FALSE
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

#install.packages("rmarkdown","pacman","tidyverse","yaml", dependencies = TRUE)
library(pacman)
pacman::p_load(ggplot2, Hmisc, tidyverse, janitor, lubridate, glue, hms,knitr, broom, ggpmisc, yaml,readxl,zoo,emmeans,lme4,multcomp)

opts_chunk$set( size="small")

# colors
sleet <- rgb(110/255, 107/255, 196/255, 1)
rain <- rgb(0/255, 0/255, 165/255, 1)
warm <- rgb(165/255, 0/255, 0/255, 1)
```

```{r}
#| label: importing-temp-data
#| echo: FALSE


data_temp <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/temperature/data/View_ClimateBasis_Nuuk_Data_Temperature_Air_temperature_@_200_cm__30min_average__DegreesC220420240157436123.csv", delim = "\t", escape_double = FALSE,col_types = cols(Date = col_datetime(format = "%Y-%m-%d"), Time = col_time(format = "%H:%M:%S")), trim_ws = TRUE) |> 
  janitor::clean_names() |> 
  drop_na() |>  
  filter(at_c != -9999) |>  
   mutate(year = year(date),
         month = month(date), 
         doy = yday(date),
         month_string = month.name[month])
```

```{r}
#| label: temp-mean-monthly
#| echo: false

data_temp_monthly_mean <- data_temp |> 
  group_by(month) |> 
  dplyr::reframe(mean = round(mean(at_c),2),
            sd = round(sd(at_c),2)) |> 
  mutate(month_string = month.name[month]) |> 
  arrange(desc(mean)) |> 
  mutate(temp_rank = row_number())

year_mean <- round(mean(data_temp$at_c),4)

data_temp_annual_mean <- data_temp |> 
  group_by(year) |> 
  filter(year != 2007) |> #excluding 2007 because it is only the months of oct, nov, dec
  dplyr::reframe(mean = mean(at_c),
            sd = sd(at_c))

```

#### Monthly mean

```{r}
#| label: fig-month-mean-temp
#| fig-scap: "Mean air temperature (°C) pr month"
#| fig-cap: !expr glue("Mean temperature pr month. Mean air temperature (°C) pr month, based on data from {min(data_temp$year)} to {max(data_temp$year)}. The dashed line represent yearly mean of {round(year_mean,3)}°C.")
#| echo: false
#| message: false
#| warning: false

ggplot(data_temp_monthly_mean, aes(x = month, y = mean))+
  geom_line(size = 0.25, linetype="solid", alpha = 0.2)+
  geom_smooth(color = "gray", se = F,  alpha = 0.1)+
  geom_point(size = 3, alpha = 0.7, color = "#076834") +
    scale_x_continuous(breaks=seq(1, 12, by = 1))+
  geom_hline(yintercept=year_mean, linetype="dotted", color = "#076834", size=0.5)+
  labs(x = "Month", y = "Mean temperature (°C)", title = "")+
  theme_minimal()+
  theme(axis.title.x = element_text(size = 8), axis.title.y = element_text(size = 8))
```

#### Yearly mean

```{r}
#| label: model-annual-mean-temp
#| 

model_temp_annual_mean <- lm(mean ~ year, data = data_temp_annual_mean)
summary(model_temp_annual_mean)

data_temp_annual_mean$pred <- predict(model_temp_annual_mean)

```

```{r}
#| label: fig-annual-mean-temp
#| fig-scap: "Mean air temperature pr year (°C)"
#| fig-cap: !expr glue("Mean annual air temperature pr year (°C) from {min(data_temp_annual_mean$year)} to {max(data_temp_annual_mean$year)} in Kangerluassunguaq. Data from 2007 is excluded because data is only from October, November and December. Gray dotted line indicate over all mean of {year_mean}°C. Dashed line is model prediction.")
#| message: false
#| warning: false
#| echo: false

ggplot(data_temp_annual_mean, aes(x = year, y = mean))+
  geom_line(size = 0.4, color = "gray", linetype="solid", alpha = 0.5)+
  #geom_smooth(method = lm, color = "#076834", se = F, linetype = "dashed")+
  geom_line(
    aes(y = pred),
    color = "#076834",
    linetype = "dashed",
  ) +
  geom_point(size = 3, alpha = 0.7, color = "#076834") +
    scale_x_continuous(breaks=seq(min(data_temp_annual_mean$year), max(data_temp_annual_mean$year), by = 1))+
  geom_hline(yintercept=year_mean, linetype="dotted", color = "darkgray", size=0.7)+
  labs(x = "Year", y = "Mean temperature (°C)", title = "")+
  theme_minimal()+
    theme(  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8),
  axis.text.x  = element_text(size = 7),
  axis.text.y  = element_text(size = 7))
```

#### Five year intervals

```{r}
#| label: data-fiveyear_intervals
#| message: false
#| warning: false
#| echo: false

data_temp_fiveyear <- data_temp |> 
  mutate(
    # choose the anchor year so that 2007–2022 is a bin
    anchor    = 2007,
    seg_start = anchor + floor((year - anchor) / 5) * 5,
    seg_end   = seg_start + 4,
    segment   = paste0(seg_start, "-", seg_end)
  ) |>
  group_by(segment, seg_start, seg_end) |>
  mutate(seg_mean_temp = mean(at_c, na.rm = TRUE))

interval_lines <- data_temp_fiveyear |>
  filter(year > 2007) |> 
  distinct(segment, seg_start, seg_end, seg_mean_temp)

```

```{r}
#| label: data-fiveyear_intervals_test
#| message: false
#| warning: false
#| echo: false

# Fit the linear mixed model
temp_data <- data_temp_fiveyear |> filter(year > 2007)

# Fit the model explicitly
model_temp_fiveyear <- lmer(
  at_c ~ segment + (1 | year), 
  data = temp_data
)

# Get estimated marginal means with cld
emm_temp_fiveyear <- model_temp_fiveyear |>
  emmeans(~ segment) |>
  cld(alpha = 0.05, Letters = letters)

cld_temp <- as.data.frame(emm_temp_fiveyear)

cld_plot <- cld_temp |>
  left_join(interval_lines, by = "segment")

# View the results
print(emm_temp_fiveyear)

```

```{r}
#| label: fig-temp-fiveyear
#| fig-scap: "5 year intervals in temperature means"
#| fig-cap: !expr glue("5 year intervals in temperature means.")
#| message: false
#| warning: false
#| echo: false

yearly_means <- data_temp_fiveyear |>
  group_by(year) |>
  filter(year > 2007) |> 
  summarise(
    mean_temp = mean(at_c, na.rm = TRUE),
    .groups = "drop"
  )

interval_lines <- data_temp_fiveyear |>
  filter(year > 2007) |> 
  distinct(segment, seg_start, seg_end, seg_mean_temp)

# amke the positions of the cld in the plot
cld_plot <- cld_plot |>
  mutate(
    x_mid = (seg_start + seg_end) / 2
  )

# 3. Plot
ggplot() +
  # Base: Yearly means (points)
  geom_hline(yintercept=year_mean, linetype="dotted", color = "darkgray", size=0.7)+
  geom_point(
    data = yearly_means,
    aes(x = year, y = mean_temp),
    size = 3,
    color = "darkgreen"
  ) +
  
  # Layer: Interval lines (horizontal spans)
  geom_segment(
    data = interval_lines,
    aes(
      x = seg_start,
      xend = seg_end,
      y = seg_mean_temp,
      yend = seg_mean_temp
    )
  ) +
  geom_text(
  data = cld_plot,
  aes(
    x = x_mid,
    y = max(yearly_means$mean_temp)*0.8,
    label = .group
  ),
  vjust = -0.8,
  size = 3
)+
  # Labels and Theme
  labs(
    x = "",
    y = "Mean Annual Temperature (°C)",
    title = ""
  ) +
  theme_minimal()+
  scale_x_continuous(breaks=seq(min(yearly_means$year), max(yearly_means$year), by = 1))+
  theme(  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8),
  axis.text.x  = element_text(size = 7),
  axis.text.y  = element_text(size = 7))
```

## Precipitation

Data is available from GEM. (2020). *ClimateBasis Nuuk—Precipitation—Precipitation accumulated (mm)* (Version 1.0) \[Dataset\]. Greenland Ecosystem Monitoring. <https://doi.org/10.17897/SXJ8-WA79>. Downloaded on XXXX.

```{r}
#| label: importing-precip
#| message: false
#| warning: false
#| echo: false

data_precip <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/precipitation/Precipitation accumulated (mm)_10.17897_SXJ8-WA79/Precipitation accumulated (mm)_10.17897_SXJ8-WA79_data.txt", delim = "\t", escape_double = FALSE, col_types = cols(Date = col_date(format = "%Y-%m-%d"), Time = col_time(format = "%H:%M:%S")), trim_ws = TRUE) |> 
  janitor::clean_names() |> 
  drop_na() |>
  filter(pre_mm != -9999) |> 
   mutate(year = year(date),
         month = month(date),
         doy = yday(date),
         month_string = month.name[month])
```
```{r}
#| label: data-precip-mean-calculations
#| message: false
#| warning: false
#| echo: false

data_precip_sum <- data_precip |> 
  group_by(year) |> 
  filter(year != 2007) |> 
  summarise(pre_mm_sum = sum(pre_mm))
data_precip_mean <- data_precip_sum |> 
  summarize(mean = mean(pre_mm_sum))

precip_year_mean <- round(mean(data_precip_mean$mean),2)

data_precip_mean_year <- data_precip  |> 
  group_by(year) |> 
  summarise(yearly_precip = sum(pre_mm, na.rm = TRUE)) |> 
  summarise(mean_precip = mean(yearly_precip, na.rm = TRUE))  |> 
  pull(mean_precip)

data_temp_hour <- data_temp |> 
  mutate(hour = substr(time, 1, 2)) |> 
  group_by(date, hour) |> 
  summarise(at_c = mean(at_c, na.rm = TRUE)) |> 
  mutate(time = hms::hms(hours = as.numeric(hour)))

sleet_lower <- -1
sleet_upper <- 1 
  
  data_pre_tem_joined <- data_precip |> 
  left_join(data_temp_hour, by = c("date" = "date", "time" = "time")) |> 
  drop_na() |> 
  mutate(precip_type = case_when(at_c < sleet_lower ~ "snow",
                              at_c <= sleet_upper ~ "sleet",
                              TRUE ~ "rain"))
  
  data_precip_month_sum <- data_precip |> 
  group_by(month, year) |> 
  summarise(pre_mm_sum = sum(pre_mm))

data_precip_month_mean <- data_precip_month_sum |> 
  group_by(month) |> 
  summarise(pre_mm_mean = round(mean(pre_mm_sum),2),
         sd = sd(pre_mm_mean)) |> 
  mutate(month_string = month.name[month]) |> 
  arrange(desc(pre_mm_mean)) |> 
  mutate(temp_rank = row_number())

precip_month_mean <- round(mean(data_precip_month_mean$pre_mm_mean),2)

data_precip_mean_annual <- data_precip  |> 
  group_by(year) |> 
  summarise(yearly_precip = sum(pre_mm, na.rm = TRUE))

```
The mean annual precipitation is `r precip_year_mean` mm.

The months with the most precipitation are `r data_precip_month_mean$month_string[1]` (`r data_precip_month_mean$pre_mm_mean[1]` mm), `r data_precip_month_mean$month_string[2]` (`r data_precip_month_mean$pre_mm_mean[2]` mm), `r data_precip_month_mean$month_string[3]` (`r data_precip_month_mean$pre_mm_mean[3]` mm), which account for `r data_precip_month_mean$pre_mm_mean[1]+data_precip_month_mean$pre_mm_mean[2]+data_precip_month_mean$pre_mm_mean[3]` mm or `r  round((data_precip_month_mean$pre_mm_mean[1]+data_precip_month_mean$pre_mm_mean[2]+data_precip_month_mean$pre_mm_mean[3])/precip_year_mean*100,1)` % of the yearly precipitation.

The months with the least precipitation are `r data_precip_month_mean$month_string[10]` (`r data_precip_month_mean$pre_mm_mean[10]` mm), `r data_precip_month_mean$month_string[11]` (`r data_precip_month_mean$pre_mm_mean[11]` mm), `r data_precip_month_mean$month_string[12]` (`r data_precip_month_mean$pre_mm_mean[12]` mm).

#### Monthly mean

```{r}
#| label: data-yearly-mean-dist_rain-snow
#| echo: false
#| message: false

# joint the temperature precipitation data to distinguish snow and rain. 

data_pre_tem_joined_monthly <- data_pre_tem_joined |> 
  group_by(year,month,precip_type) |> 
  summarise(sum = sum(pre_mm)) |> 
  filter(year != 2007)

data_pre_tem_monthly <- data_pre_tem_joined_monthly |> 
  group_by(month,precip_type) |> 
  summarise(mean = round(mean(sum),2)) 

data_pre_tem_year <- data_pre_tem_joined_monthly |> 
  group_by(year,precip_type) |> 
  summarise(mean = round(sum(sum),2)) 
```

```{r}
#| label: fig-snow-rain-year
#| echo: false
#| message: false
#| warning: false
#| fig-scap: "Mean monthly distribution of types of precipitation (mm)"
#| fig-cap: !expr glue("Mean precipitation type pr month. Mean monthly distribution of types of precipitation (mm). Based on data from {min(data_pre_tem_joined_monthly$year)} to {max(data_pre_tem_joined_monthly$year)}. Sleet is defined as precipitation that fell when temperatures were between {sleet_lower}°C and {sleet_upper}°C.")

ggplot(data_pre_tem_monthly, aes(x = month, y = mean, fill = precip_type)) + 
  geom_bar(stat = "identity")+
  scale_x_continuous(breaks=seq(min(data_pre_tem_monthly$month), max(data_pre_tem_monthly$month), by = 1))+
  labs(x = "Month", y = "Mean precipitation (mm)", title = "")+
  scale_fill_manual(values = c(rain,sleet,"#b3b3b3"))+
  guides(fill = guide_legend(title = ""))+
  theme_minimal()+
      theme(  axis.title.x = element_text(size = 8),
  axis.title.y = element_text(size = 8),
  axis.text.x  = element_text(size = 7),
  axis.text.y  = element_text(size = 7))

```

#### Yearly mean

```{r}
#| label: model-annual-mean-precip
#| echo: false

model_precip_mean_annual <- lm(yearly_precip ~ year, data = data_precip_mean_annual)
summary(model_precip_mean_annual)

data_precip_mean_annual$pred <- predict(model_precip_mean_annual)

```

```{r}
#| label: fig-precipitation-annual
#| echo: false
#| fig-scap: "Mean yearly precipitation (mm)"
#| fig-cap: !expr glue("Mean yearly precipitation (mm) from {min(data_precip_sum$year)} to {max(data_precip_sum$year)}. Data from 2007 is excluded because it was only from May - December. Blue dashed line indicate over all mean of {precip_year_mean} mm.")
#| message: false
#| warning: false

ggplot(data_precip_sum, aes(y = pre_mm_sum, x = year))+
  geom_bar(stat = "identity", fill = "#6e6bc4")+
  scale_x_continuous(breaks=seq(min(data_precip_sum$year), max(data_precip_sum$year), by = 1))+
geom_hline(yintercept = precip_year_mean, linetype="dotted", color = "blue", linewidth =0.5)+
  geom_smooth(method = lm, color = "blue", size = 0.5, se = F, linetype="dashed")+
  labs(x = "Year", y = "Precipitation (mm)")+
  stat_poly_eq(
      aes(label = paste(..eq.label.., ..rr.label.., ..p.value.label.., sep = "~~~")),
      formula = y ~ x,
      parse = TRUE
    )+
  theme_minimal()

```

#### Five year intervals

# Historic climate data

## Temperature

## Precipitation

# Statistical analysis

## Diversity

### Richness

### Evenness

### Shannon

### Turnover

## Community composition

### Within vegetation composition

### Between vegetation composition

## Shrubification

### Functional groups

### Specific species
