---
title: "macroclimate"
author: "ibdj"
date: last-modified
date-format: "YYYY-MM-DD HH:mm"
format:
  pdf:
    fig-pos: "H"
    fig-width: 9
    fig-height: 5
    documentclass: scrartcl
    toc: true
    lof: true
    number-sections: true
    colorlinks: true
editor: visual
execute: 
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

# Macroclimate

Background macroclimate of Kangerluarsunnguaq ([Kobbefjord, Nuuk 64.13334, -51.34368](https://maps.app.goo.gl/2ereBF15bAWeKRVX9)). Data produced by [Asiaq - Greenland Survey](https://www.asiaq.gl/en/) for [Greenland Ecosystem Monitoring](https://data.g-e-m.dk) [@asiaqa; @asiaq; @asiaqb].

```{r}
#| label: importing-packages
#| output: FALSE

install.packages("rmarkdown","pacman","tidyverse")
library(pacman)
pacman::p_load(tidyverse, janitor, lubridate, glue, hms,knitr, broom)

opts_chunk$set( size="small")

# colors
sleet <- rgb(110/255, 107/255, 196/255, 1)
rain <- rgb(0/255, 0/255, 165/255, 1)
warm <- rgb(165/255, 0/255, 0/255, 1)
```

# Temperature data

```{r}
#| label: importing-temp-data
#| 
data_temp <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/temperature/data/View_ClimateBasis_Nuuk_Data_Temperature_Air_temperature_@_200_cm__30min_average__DegreesC220420240157436123.csv", delim = "\t", escape_double = FALSE,col_types = cols(Date = col_datetime(format = "%Y-%m-%d"), Time = col_time(format = "%H:%M:%S")), trim_ws = TRUE) |> 
  janitor::clean_names() |> 
  drop_na() |>  
  filter(at_c != -9999) |>  
   mutate(year = year(date),
         month = month(date), 
         doy = yday(date),
         month_string = month.name[month])
```

```{r}
#| label: temp-mean-monthly

data_temp_monthly_mean <- data_temp |> 
  group_by(month) |> 
  summarize(mean = round(mean(at_c),2),
            sd = round(sd(at_c),2)) |> 
  mutate(month_string = month.name[month]) |> 
  arrange(desc(mean)) |> 
  mutate(temp_rank = row_number())

```

```{r}
#| label: temp-mean-annual

data_temp_annual_mean <- data_temp |> 
  group_by(year) |> 
  filter(year != 2007) |> #excluding 2007 because it is only the months of oct, nov, dec
  summarize(mean = mean(at_c),
            sd = sd(at_c))
```

The imported data has entries between `r min(data_temp$year)` and `r max(data_temp$year)`.

Temperatures between `r min(data_temp$at_c)`°C (`r data_temp$date[which.min(data_temp$at_c)]`) and `r max(data_temp$at_c)`°C (`r data_temp$date[which.max(data_temp$at_c)]`) have been recorded.

## Mean annual temperature

```{r}
#| label: caluclating-year-mean
#| echo: false

year_mean <- round(mean(data_temp$at_c),4)

```

The mean temperature across all entries is `r year_mean`°C.

The warmest months are `r data_temp_monthly_mean$month_string[1]` (`r data_temp_monthly_mean$mean[1]`°C), `r data_temp_monthly_mean$month_string[2]` (`r data_temp_monthly_mean$mean[2]`°C) and `r data_temp_monthly_mean$month_string[3]` (`r data_temp_monthly_mean$mean[3]`°C).

The coldest months are `r data_temp_monthly_mean$month_string[12]` (`r data_temp_monthly_mean$mean[12]`°C), `r data_temp_monthly_mean$month_string[11]` (`r data_temp_monthly_mean$mean[11]`°C), and `r data_temp_monthly_mean$month_string[10]` (`r data_temp_monthly_mean$mean[10]`°C).

```{r}
#| label: fig-month-mean-temp
#| fig-scap: "Mean air temperature (°C) pr month"
#| fig-cap: !expr glue("Mean air temperature (°C) pr month, based on data from {min(data_temp$year)} to {max(data_temp$year)}. The dashed line represent yearly mean of {round(year_mean,3)}°C.")
#| message: false
#| warning: false

ggplot(data_temp_monthly_mean, aes(x = month, y = mean))+
  geom_smooth(color = "blue", se = F)+
  geom_point()+
  geom_line(size = 0.25, linetype="dotted")+
  scale_x_continuous(breaks=seq(1, 12, by = 1))+
  geom_hline(yintercept=year_mean, linetype="dashed", color = "blue", size=0.5)+
  labs(x = "Month", y = "Mean temperature (°C)", title = "Mean temperature pr month")
```



```{r}
#| label: fig-annual-mean-temp
#| fig-scap: "Mean air temperature pr year (°C)"
#| fig-cap: !expr glue("Mean air temperature pr year (°C) from {min(data_temp_annual_mean$year)} to {max(data_temp_annual_mean$year)}. Data from 2007 is excluded because data is only from October, November and December. Blue dashed line indicate over all mean of {year_mean}°C. Solid blue is trend line.")
#| message: false
#| warning: false

ggplot(data_temp_annual_mean, aes(x = year, y = mean))+
  geom_smooth(method = lm, color = "blue", se = F)+
  geom_point(size = 2)+
  geom_line(size = 0.25, color = "black", linetype="dotted")+
  scale_x_continuous(breaks=seq(min(data_temp_annual_mean$year), max(data_temp_annual_mean$year), by = 1))+
  geom_hline(yintercept=year_mean, linetype="dashed", color = "blue", size=0.5)+
  labs(x = "Year", y = "Mean temperature (°C)", title = "Annual mean temperature, Kangerluassunguaq")
```
{{< pagebreak >}}
## Mean temperature of warmest months

```{r}
#| label: mean-temp-warmest-months
#| output: FALSE
#| message: false
#| warning: false

data_temp_monthly_mean_summer <- data_temp |> 
  filter(month %in% c(6,7,8)) |> 
  group_by(month, year) |> 
  summarize(mean = mean(at_c))

data_temp_mean_warmest <- mean(data_temp_monthly_mean_summer$mean)
```

```{r}
#| label: linear-regression-varmest-months
#| output: false
#| message: false
#| warning: false

model_temp_warmest <- lm(mean ~ year, data = data_temp_monthly_mean_summer)

summary(model_temp_warmest)

model_temp_warmest_p_value <- glance(model_temp_warmest)$p.value
```

```{r}
#| label: fig-month-summer-mean-temp
#| fig-scap: "Mean air temperature pr year (°C) for warmest months"
#| fig-cap: !expr glue("Mean air temperature (°C) pr for {data_temp_monthly_mean$month_string[1]}, {data_temp_monthly_mean$month_string[2]}, and {data_temp_monthly_mean$month_string[3]} based on data from {min(data_temp$year)} to {max(data_temp$year)}. P-value for the regression i {round(model_temp_warmest_p_value,3)}. The mean of the warmests months across all years are {round(data_temp_mean_warmest,3)}°C.")
#| message: false
#| warning: false

ggplot(data_temp_monthly_mean_summer, aes(x = year, y = mean, color = as.factor(month)))+
  geom_smooth(method = lm, color = "gray", se = F, size = 1, linetype = 2)+
  geom_point(size = 2)+
  scale_x_continuous(breaks=seq(min(data_temp_monthly_mean_summer$year), max(data_temp_monthly_mean_summer$year), by = 1))+
  #geom_hline(yintercept=year_mean, linetype="dashed", color = "blue", size=0.5)+
  labs(x = "Month", y = "Mean temperature (°C)", title = "Mean temperature pr summer month")+
  labs(color = "Months")
```

```{r}
#| label: regression-warmest-temp-months

model_temp_warmest <- lm(mean ~ year, data = data_temp_monthly_mean_summer)

summary(model_temp_warmest)

model_temp_warmest_p_value <- glance(model_temp_warmest)$p.value
```

## Temperature anomolies
```{r}
#| label: data-monthly-temp-anomaly
# Data prepping for temperature anomalies heat map

data_temp_monthly_mean <- data_temp |> 
  group_by(year, month) |> 
  summarise(monthly_mean = mean(at_c, na.rm = TRUE)) |> 
  ungroup()

data_temp_monthly_mean_overall <- data_temp_monthly_mean |> 
  group_by(month) |> 
  summarise(overall_monthly_mean = mean(monthly_mean, na.rm = TRUE)) |> 
  ungroup()

data_temp_anomaly <- data_temp_monthly_mean |> 
  left_join(data_temp_monthly_mean_overall, by = "month") |> 
  mutate(temperature_anomaly = monthly_mean - overall_monthly_mean)

```

```{r}
#| label: fig-monthly-temp-anomaly
#| fig-scap: "Monthly Temperature Anomalies Heatmap"
#| fig-cap: !expr glue("Monthly Temperature Anomalies Heatmap. The stronger the color the bigger the difference from mean temperature of the respective month across all monitored years (°C).")


ggplot(data_temp_anomaly, aes(x = factor(year), y = factor(month), fill = temperature_anomaly)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = rain, high = warm, mid = "white", 
                       midpoint = 0, limit = c(-max(abs(data_temp_anomaly$temperature_anomaly)), 
                                               max(abs(data_temp_anomaly$temperature_anomaly))),
                       space = "Lab", name="Temperature Anomaly (°C)") +
  labs(x = "Year", y = "Month", title = "Monthly Temperature Anomalies Heatmap") +
  scale_y_discrete(labels = month.name) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "right",
    legend.direction = "vertical",
    legend.title = element_text(angle = 90, hjust = 0.5, vjust = 1, size = 10),
    legend.text = element_text(angle = 90, hjust = 1),
    legend.key.height = unit(1, "cm"),
    legend.key.width = unit(0.5, "cm"),
    legend.box.just = "bottom") +
 guides(fill = guide_colorbar(title.position = "left", label.position = "left"))


```

## Absolute values of anomolies

```{r}
#| label: data-absolut-abnormalities-temp
#| message: false
#| warning: false

data_temp_anomaly_abs <- data_temp_anomaly|> 
  mutate(abs_anomo = abs(temperature_anomaly))

```

```{r}
#| label: fig-temp-abs-anomolies
#| fig-scap: "Absolut values of anomolies of PAR"
#| fig-cap: !expr glue("Absolut values of anomolies (difference from mean) of respective month.")
#| message: false
#| warning: false

ggplot(data_temp_anomaly_abs, aes(x = year, y = abs_anomo)) +
  geom_point() +
  geom_smooth(method = "lm")+
  xlab("Year") +
  ylab("Absolute anomolies, temp") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_x_continuous(breaks=seq(min(data_temp_anomaly_abs$year), max(data_temp_anomaly_abs$year), by = 1))+
  theme_minimal()
```


# Precipitation data

```{r}
#| label: importing-precip

data_precip <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/precipitation/data/View_ClimateBasis_Nuuk_Data_Precipitation_Precipitation_accumulated_mm210420242020078158.csv", delim = "\t", escape_double = FALSE, col_types = cols(Date = col_date(format = "%Y-%m-%d"), Time = col_time(format = "%H:%M:%S")), trim_ws = TRUE) |> 
  janitor::clean_names() |> 
  drop_na() |>
  filter(pre_mm != -9999) |> 
   mutate(year = year(date),
         month = month(date),
         doy = yday(date),
         month_string = month.name[month])
```

## Yearly precipitation

```{r}
#| label: precip-mean-annual

data_precip_sum <- data_precip |> 
  group_by(year) |> 
  filter(year != 2007) |> 
  summarise(pre_mm_sum = sum(pre_mm))
```

```{r}
#| label: precip-year-mean
#| echo: false

data_precip_mean <- data_precip_sum |> 
  summarize(mean = mean(pre_mm_sum))

precip_year_mean <- round(mean(data_precip_mean$mean),2)
```

The mean annual precipitation is `r precip_year_mean` mm.

```{r}
#| label: fig-precipitation-annual
#| fig-scap: "Mean yearly precipitation (mm)"
#| fig-cap: !expr glue("Mean yearly precipitation (mm) from {min(data_precip_sum$year)} to {max(data_precip_sum$year)}. Data from 2007 is excluded because it was only from May - December. Blue dashed line indicate over all mean of {precip_year_mean} mm.")

ggplot(data_precip_sum, aes(y = pre_mm_sum, x = year))+
  geom_bar(stat = "identity", fill = "#6e6bc4")+
  scale_x_continuous(breaks=seq(min(data_precip_sum$year), max(data_precip_sum$year), by = 1))+
geom_hline(yintercept = precip_year_mean, linetype="dashed", color = "blue", linewidth =0.5)+
  geom_smooth(method = lm, color = "blue", size = 0.5, se = F)+
  labs(x = "Year", y = "Precipitation (mm)")

```

## Monthly precipitation

```{r}
#| label: precip-mean-monthly
#| warning: false

data_precip_month_sum <- data_precip |> 
  group_by(month, year) |> 
  summarise(pre_mm_sum = sum(pre_mm))

data_precip_month_mean <- data_precip_month_sum |> 
  group_by(month) |> 
  summarise(pre_mm_mean = round(mean(pre_mm_sum),2),
         sd = sd(pre_mm_mean)) |> 
  mutate(month_string = month.name[month]) |> 
  arrange(desc(pre_mm_mean)) |> 
  mutate(temp_rank = row_number())

precip_month_mean <- round(mean(data_precip_month_mean$pre_mm_mean),2)

```

```{r}
#| label: fig-precipitation-month
#| fig-scap: "Mean monthly precipitation (mm)"
#| fig-cap: !expr glue("Mean monthly precipitation (mm) including data from {min(data_precip$year)} to {max(data_precip$year)}. Blue dashed line represent monthly mean of {precip_month_mean} mm. ADD SF BARS!")
#| warning: false

ggplot(data_precip_month_mean, aes(y = pre_mm_mean, x = month))+
  geom_smooth(color = "blue", se = F)+
  geom_bar(stat = "identity", fill = "#6e6bc4", alpha = 0.95)+
  scale_x_continuous(breaks=seq(min(data_precip_month_mean$month), max(data_precip_month_mean$month), by = 1))+
geom_hline(yintercept = precip_month_mean, linetype="dashed", color = "blue", linewidth =0.25)
```

The months with the most precipitation are `r data_precip_month_mean$month_string[1]` (`r data_precip_month_mean$pre_mm_mean[1]` mm), `r data_precip_month_mean$month_string[2]` (`r data_precip_month_mean$pre_mm_mean[2]` mm), `r data_precip_month_mean$month_string[3]` (`r data_precip_month_mean$pre_mm_mean[3]` mm).

The months with the least precipitation are `r data_precip_month_mean$month_string[10]` (`r data_precip_month_mean$pre_mm_mean[10]` mm), `r data_precip_month_mean$month_string[11]` (`r data_precip_month_mean$pre_mm_mean[11]` mm), `r data_precip_month_mean$month_string[12]` (`r data_precip_month_mean$pre_mm_mean[12]` mm).

## Precipitation type

```{r}
#| label: temp-data-hourly

# making the temperature data hourly 

data_temp_hour <- data_temp |> 
  mutate(hour = substr(time, 1, 2)) |> 
  group_by(date, hour) |> 
  summarise(at_c = mean(at_c, na.rm = TRUE)) |> 
  mutate(time = hms::hms(hours = as.numeric(hour)))

```

```{r}
#| label: joining-temp-precip-data

# joint the temperature precipitation data to destinguish snow and rain. 

# data_pre_tem_joined <- data_precip |> 
#   left_join(data_temp_hour, by = c("date" = "date", "time" = "time")) |> 
#   drop_na() |> 
#   mutate(precip_type = ifelse(at_c < 0,"snow","rain"))

sleet_lower <- -1
sleet_upper <- 1 
  
  data_pre_tem_joined <- data_precip |> 
  left_join(data_temp_hour, by = c("date" = "date", "time" = "time")) |> 
  drop_na() |> 
  mutate(precip_type = case_when(at_c < sleet_lower ~ "snow",
                              at_c <= sleet_upper ~ "sleet",
                              TRUE ~ "rain"))
```

```{r}
#| label: monthly-dist-rain-snow

# joint the temperature precipitation data to destinguish snow and rain. 

data_pre_tem_joined_monthly <- data_pre_tem_joined |> 
  group_by(year,month,precip_type) |> 
  summarise(sum = sum(pre_mm)) |> 
  filter(year != 2007)
```

```{r}
#| label: monthly-mean-dist_rain-snow

# joint the temperature precipitation data to destinguish snow and rain. 

data_pre_tem_monthly <- data_pre_tem_joined_monthly |> 
  group_by(month,precip_type) |> 
  summarise(mean = round(mean(sum),2)) 
```

```{r}
#| label: yearly-mean-dist_rain-snow

# joint the temperature precipitation data to distinguish snow and rain. 

data_pre_tem_year <- data_pre_tem_joined_monthly |> 
  group_by(year,precip_type) |> 
  summarise(mean = round(sum(sum),2)) 
```

```{r}
#| label: fig-snow-rain-year
#| fig-scap: "Mean monthly distribution of types of precipitation (mm)"
#| fig-cap: !expr glue("Mean monthly distribution of types of precipitation (mm). Based on data from {min(data_precip$year)} to {max(data_precip$year)}. Sleet is defined as precipitation that fell when temperatures were between {sleet_lower}°C and {sleet_upper}°C.")

ggplot(data_pre_tem_monthly, aes(x = month, y = mean, fill = precip_type)) + 
  geom_bar(stat = "identity")+
  scale_x_continuous(breaks=seq(min(data_pre_tem_monthly$month), max(data_pre_tem_monthly$month), by = 1))+
  labs(x = "Month", y = "Mean precipitation (mm)", title = "Mean precipitation type pr month")+
  scale_fill_manual(values = c(rain,sleet,"#b3b3b3"))+
  guides(fill = guide_legend(title = ""))

```

```{r}
#| label: fig-snow-rain
#| fig-scap: "Mean percipitation pr. month"
#| fig-cap: !expr glue("Mean percipitation pr. month over the monitoring period ({min(data_pre_tem_year$year)} - {max(data_pre_tem_year$year)}). Sleet is defined as precipitation that fell when temperatures were between {sleet_lower}°C and {sleet_upper}°C.")

ggplot(data_pre_tem_year, aes(x = year, y = mean, fill = precip_type)) + 
  geom_bar(stat = "identity")+
  scale_x_continuous(breaks=seq(min(data_pre_tem_year$year), max(data_pre_tem_year$year), by = 1))+
  labs(x = "Year", y = "Mean precipitation (mm)", title = "Mean precipitation type pr year")+
  scale_fill_manual(values = c(rain,sleet ,"#b3b3b3"))+
  geom_smooth(aes(color = precip_type), method = "lm", se = FALSE, show.legend = FALSE, size = 0.5) + 
  scale_color_manual(values = c(rain, sleet, "#b3b3b3")) +
  guides(fill = guide_legend(title = ""))

```

## Precipitation anomalies

```{r}
#| label: data-precip-anomalies

# Data prepping for precip anomalies heat map

data_precip_month_sum <- data_precip |> 
  group_by(year, month) |> 
  summarise(monthly_pre_sum = sum(pre_mm, na.rm = TRUE)) |> 
  ungroup()

data_precip_monthly_mean_sum <- data_precip_month_sum |> 
  group_by(month) |> 
  summarise(overall_pre_monthly_mean = mean(monthly_pre_sum, na.rm = TRUE)) |> 
  ungroup()

monthly_prep_anomalies <- data_precip_monthly_mean_sum |> 
  left_join(data_precip_month_sum, by = "month") |> 
  mutate(prep_anomaly = monthly_pre_sum - overall_pre_monthly_mean)

```

```{r}
#| label: fig-monthly-prep-anomaly
#| fig-scap: "Monthly precipitation anomaly (°C)"
#| fig-cap: !expr glue("Monthly precipitation anomaly (°C). Stronger colors indicate large derivation from the month mean.")

ggplot(monthly_prep_anomalies, aes(x = factor(year), y = factor(month), fill = prep_anomaly)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "orange", high = rain, mid = "white", 
                       midpoint = 0, limit = c(-max(abs(monthly_prep_anomalies$prep_anomaly)), 
                                               max(abs(monthly_prep_anomalies$prep_anomaly))),
                       space = "Lab", name="Precipitation Anomaly (mm)") +
  labs(x = "Year", y = "Month", title = "Monthly Precipitation Anomalies Heatmap") +
  scale_y_discrete(labels = month.name) + # Use full month names on y-axis
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "right",
    legend.direction = "vertical",
    legend.title = element_text(angle = 90, hjust = 0.5, vjust = 1, size = 10),
    legend.text = element_text(angle = 90, hjust = 1),
    legend.key.height = unit(1, "cm"),
    legend.key.width = unit(0.5, "cm"),
    legend.box.just = "bottom") +
 guides(fill = guide_colorbar(title.position = "left", label.position = "left"))

```

## Absolute values of anomolies

```{r}
#| label: data-absolut-abnormalities-prep
#| message: false
#| warning: false

data_prep_anomaly_abs <- monthly_prep_anomalies|> 
  mutate(abs_anomo = abs(prep_anomaly))

```

```{r}
#| label: fig-prep-abs-anomolies
#| fig-scap: "Absolut values of anomolies of PAR"
#| fig-cap: !expr glue("Absolut values of anomolies (difference from mean) of respective month.")
#| message: false
#| warning: false

ggplot(data_prep_anomaly_abs, aes(x = year, y = abs_anomo)) +
  geom_point() +
  geom_smooth(method = "lm")+
  xlab("Year") +
  ylab("Absolute anomolies, prep") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_x_continuous(breaks=seq(min(data_prep_anomaly_abs$year), max(data_prep_anomaly_abs$year), by = 1))+
  theme_minimal()
```

# PAR data

```{r}
#| label: data-par-importing


par <- View_ClimateBasis_Nuuk_Data_Radiation_Photosyntetic_active_radiation_200_cm_5min_average_æmol_m2_sec210420241337080578 <- read_delim("~/Library/CloudStorage/OneDrive-Aarhusuniversitet/MappingPlants/01 Vegetation changes Kobbefjord/data/r_climatedata/data/ration_par/View_ClimateBasis_Nuuk_Data_Radiation_Photosyntetic_active_radiation_@_200_cm__5min_average_æmol_m2_sec210420241337080578.csv", delim = "\t", escape_double = FALSE, trim_ws = TRUE) |> 
  clean_names() |> 
  filter(par_mmol_sec_m2 != -9999,
         solar_elevation > 0 ) |> 
  mutate(year = year(date),
         month = factor(month(date, label = TRUE, abbr = FALSE), levels = month.name))
```

## Monthly means

```{r}
#| label: data-par-mean-monthly
par_monthly_mean <- View_ClimateBasis_Nuuk_Data_Radiation_Photosyntetic_active_radiation_200_cm_5min_average_æmol_m2_sec210420241337080578 |> 
  filter(solar_elevation > 0) |> 
  group_by(month = lubridate::month(date, label = TRUE, abbr = FALSE)) |>  # Group by month name (e.g., "Jan")
  summarize(par_monthly_mean = mean(par_mmol_sec_m2, na.rm = TRUE)) |>  # Calculate mean PAR
  arrange(desc(par_monthly_mean)) |>  # Sort descending by mean PAR
  mutate(temp_rank = row_number())   # Add ranks

```

```{r}
#| label: fig-par-monthly-mean
#| fig-scap: "Monthly mean PAR"
#| fig-cap: !expr glue("Monthly mean PAR from {min(par$year)} to {max(par$year)}. Calculations based on PAR values only when the sun is up (solar elevation > 0).")
#| message: false
#| warning: false

# Plot
ggplot(par, aes(x = month, y = par_mmol_sec_m2)) +
  geom_boxplot() +
  xlab("Month") +
  ylab("PAR (mmol/sec/m2)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Highest monthly means

The months with the highest mean levels of PAR (when the sun is up, solar elevation \> 0) are `r par_monthly_mean$month[1]` (`r round(par_monthly_mean$par_monthly_mean[1],1)`), `r par_monthly_mean$month[2]` (`r round(par_monthly_mean$par_monthly_mean[2],1)`), `r par_monthly_mean$month[3]` (`r round(par_monthly_mean$par_monthly_mean[3],1)`).

```{r}
#| label: par-highest-mean

par_highest_mean <- par |> 
  group_by(year, month) |>  # Group by month name (e.g., "Jan")
  filter(month == c("May","June","July")) |> 
  summarize(par_mean = mean(par_mmol_sec_m2, na.rm = TRUE)) |>  # Calculate mean PAR
  arrange(desc(par_mean)) |>  # Sort descending by mean PAR
  mutate(temp_rank = row_number())   # Add ranks
```

```{r}
#| label: fig-par-month-highest-means
#| fig-scap: "Mean of months with highest PAR"
#| fig-cap: !expr glue("Mean of months with highest PAR.")
# Plot
par_highest_mean <- par_highest_mean %>%
  mutate(month_num = as.numeric(month))

ggplot(par_highest_mean, aes(x = month_num, y = par_mean)) +
  geom_point(aes(color = factor(year)), size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgray", linetype = "dashed") +
  scale_x_continuous(
    breaks = 1:12,
    labels = levels(par_highest_mean$month)
  ) +
  xlab("Month") +
  ylab("PAR (mmol/sec/m2)") +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  labs(color = "Year")
```
## Observations above 2000
```{r}
#| label: fig-par-high-values
#| fig-scap: "Number of measurements of PAR above 2000"
#| fig-cap: !expr glue("Number of measurements of PAR above 2000.")
# measurements above mean
# Step 1: Calculate the monthly mean and add it to the dataframe
above_2000 <- par |> 
  group_by(month,year) |> 
  filter(par_mmol_sec_m2 > 2000) |> 
  summarize(above_2000 = n())

# Step 4: Plot
ggplot(above_2000, aes(x = year, y = above_2000)) +
  geom_point(color = "darkgreen", alpha = 0.8)+
  geom_smooth(method = "lm", linetype = "dashed")+
  xlab("Year") +
  ylab("Number of measurements above 2000")

# Fit the model
model_above <- lm(above_2000 ~ year, data = above_2000)

# View the summary (includes p-value)
summary(model_above)

```

```{r}
#| label: regression-par-month
# Convert month to numeric if necessary

# Fit the model
model <- lm(par_mean ~ month_num, data = par_highest_mean)

# View the summary (includes p-value)
summary(model)
```

## PAR anomalies

```{r}
#| label: fig-par-anomolies-heatmap
#| fig-scap: "Monthly PAR anomaly"
#| fig-cap: !expr glue("Anomolies (difference from mean) of respective month. Stronger color indication bigger difference from monthly mean across entire period.")
#| message: false
#| warning: false

par_monthly_anomalies <- par |>
  group_by(month) |> 
  mutate(month_mean = mean(par_mmol_sec_m2, na.rm = TRUE)) |> 
  ungroup() |> 
  group_by(year, month) |> 
  summarise(
    mean = mean(par_mmol_sec_m2, na.rm = TRUE),
    month_mean = mean(month_mean, na.rm = TRUE), # will be the same for all rows in a month
    anomaly = mean - mean(month_mean, na.rm = TRUE)
  ) |> 
  ungroup()

ggplot(par_monthly_anomalies, aes(x = factor(year), y = factor(month), fill = anomaly)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = rain, high = warm, mid = "white", 
                       midpoint = 0, limit = c(-max(abs(par_monthly_anomalies$anomaly)), 
                                               max(abs(par_monthly_anomalies$anomaly))),
                       space = "Lab", name="PAR Anomaly") +
  labs(x = "", y = "", title = "PAR Anomalies Heatmap") +
  scale_y_discrete(labels = month.name) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "right",
    legend.direction = "vertical",
    legend.title = element_text(angle = 90, hjust = 0.5, vjust = 1, size = 10),
    legend.text = element_text(angle = 90, hjust = 1),
    legend.key.height = unit(1, "cm"),
    legend.key.width = unit(0.5, "cm"),
    legend.box.just = "bottom") +
 guides(fill = guide_colorbar(title.position = "left", label.position = "left"))

```

## Absolute values of anomolies

```{r}
#| label: data-absolut-abnormalities-par
#| message: false
#| warning: false

par_anomalities_abs <- par_monthly_anomalies |> 
  mutate(abs_anomo = abs(anomaly))

```

```{r}
#| label: fig-par-abs-anomolies
#| fig-scap: "Absolut values of anomolies of PAR"
#| fig-cap: !expr glue("Absolut values of anomolies (difference from mean) of respective month.")
#| message: false
#| warning: false

ggplot(par_anomalities_abs, aes(x = year, y = abs_anomo)) +
  geom_point() +
  geom_smooth(method = "lm")+
  xlab("Year") +
  ylab("Absolute anomolies, PAR (mmol/sec/m2)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme_minimal()
```


# Historic climate context

Lookings to see if CHELSA data shoc the same pattern of minimal change as the local data. Using this guide to optain data: https://gitlabext.wsl.ch/karger/rchelsa

```{r}
#| label: data-global-climate-data-import

# # Install 'rchelsa' in R:
# # https://github.com/inSileco/rchelsa
# install.packages("remotes")
# remotes::install_github("inSileco/rchelsa")

```


# References
